% ==============================================================================
% Section: Calibration
% Status: PARTIAL (2026-03-27) -- data sources + VRR parameters complete;
%         market structure, cost parameters, delivery years TBD (Session 2+)
% ==============================================================================

\section{Calibration}
\label{sec:calibration}

% ------------------------------------------------------------------------------
\subsection{Data Sources}
\label{subsec:cal_data}
% ------------------------------------------------------------------------------

The calibration draws on four publicly available datasets produced by PJM
and the PJM Independent Market Monitor (IMM).

\paragraph{PJM Planning Parameters.}
PJM publishes the planning parameters for each BRA in advance of the
auction, including the VRR curve anchor points, Net CONE, CETL, and
reliability requirement for every LDA. These files are available in Excel
format and provide the complete VRR demand schedule that sellers observe
when submitting capacity offers. We processed the planning parameter files
for delivery years 2021/22 through 2028/29 using a custom parser that
extracts LDA-level parameters for each VRR design vintage and handles the
three-point (old) and four-point (new) curve specifications separately.
Spot-checks against published PJM summary tables confirm accuracy (e.g.,
2023/24 RTO Net CONE = \$274.96 per MW-day).

\paragraph{PJM BRA Results.}
Following each auction, PJM releases clearing prices and quantities by LDA.
We processed seven completed auctions (2021/22 through 2027/28); the
2028/29 BRA had not yet been conducted at the time of data collection.
Clearing results are verified against PJM press releases (e.g., 2025/26
RTO clearing = \$269.92 per MW-day; 2026/27 RTO clearing = \$329.17).

\paragraph{PJM IMM State of the Market Reports.}
The PJM IMM publishes an annual State of the Market (SotM) report containing
auction-by-auction summaries of market concentration, TPS test outcomes, and
offer-curve characteristics. We extracted Residual Supply Index values and
participant counts from the relevant report volumes using a PDF-parsing
pipeline. Volume 2022 covers delivery years through 2022/23, and Volume 2025
covers delivery years 2023/24 through 2027/28. Due to layout variation
across report vintages, RSI data is available for a subset of LDAs and
delivery years, as documented in Table~\ref{tab:tps_results}. The 2021/22
delivery year predates our RSI extraction window and is excluded from the
market structure analysis.

\paragraph{Auction Lead Times.}
BRA settlement dates and delivery-year start dates were coded manually from
PJM auction notices and verified against the data in
Table~\ref{tab:bra_timeline}. Lead time is defined as the number of calendar
months between the BRA settlement date and the first day of the delivery
year. The highly irregular lead-time pattern documented in
Section~\ref{subsec:inst_leadtime} is incorporated as a covariate in the
calibration of the investment-signal environment.

The four datasets are joined at the LDA--delivery-year level to produce a
master panel with 110 observations spanning 17 LDAs and seven delivery years
(2021/22 through 2027/28).

% ------------------------------------------------------------------------------
\subsection{VRR Curve Parameters}
\label{subsec:cal_vrr}
% ------------------------------------------------------------------------------

Table~\ref{tab:vrr_params} reports the VRR curve parameters for the RTO
under both the old and new designs for the six delivery years with complete
BRA data. Under the old design, Point~(a) equals $1.5 \times \text{Net CONE}$
and Point~(b) equals $0.75 \times \text{Net CONE}$ by construction, so the
ratio $p_{a} / p_{b} = 2$ is exact for all old-design years except 2025/26,
where a transitional administrative cap produced a Point~(a) price above the
standard $1.5 \times \text{Net CONE}$ multiple.\footnote{The 2025/26 Net CONE
of \$228.81 per MW-day implies a standard cap of \$343.22, but the published
Point~(a) was \$451.61. This discrepancy reflects the fact that VRR anchor
points are fixed at planning-parameter publication and may diverge from the
final Net CONE if energy and ancillary service revenues are revised
substantially between publication and auction. We use the published anchor
points as the realized VRR parameters in all calibrations.}
Under the new design, Points~(b) and~(c) equal the explicit floor price of
approximately \$177--\$179 per MW-day, and Point~(a) is set at approximately
$1.5 \times \text{Net CONE}$.

\begin{table}[H]
  \centering
  \caption{RTO VRR Curve Parameters, 2022/23--2027/28}
  \label{tab:vrr_params}
  \begin{tabular}{lcccccc}
    \toprule
    Delivery & Net CONE    & Pt.~(a) & Pt.~(a) & Pt.~(b) & Pt.~(b) & VRR  \\
    Year     & (\$/MW-day) & Price   & MW      & Price   & MW      & Design \\
    \midrule
    2022/23 & \$260.50 & \$390.75 & 136{,}076 & \$195.38 & 139{,}656 & Old \\
    2023/24 & \$274.96 & \$412.44 & 135{,}914 & \$206.22 & 139{,}473 & Old \\
    2024/25 & \$293.19 & \$439.79 & 138{,}341 & \$219.89 & 141{,}910 & Old \\
    2025/26 & \$228.81 & \$451.61 & 133{,}554 & \$171.61 & 137{,}160 & Old \\
    2026/27 & \$212.14 & \$329.17 & 134{,}887 & \$177.24 & 136{,}361 & New \\
    2027/28 & \$242.52 & \$333.44 & 141{,}681 & \$179.55 & 143{,}218 & New \\
    \bottomrule
  \end{tabular}
  \vspace{3pt}

  \begin{minipage}{0.92\textwidth}
    \footnotesize \textit{Note:} Prices in nominal dollars per MW-day;
    quantities in MW. Under the old design, Point~(c) is the zero-price
    floor at a quantity above Point~(b); under the new design, Points~(b)
    and~(c) share the floor price and a flat segment extends to Point~(d).
    The 2025/26 Point~(a) price exceeds $1.5 \times \text{Net CONE}$ due to
    a transitional administrative cap (see footnote above). Data source: PJM
    Planning Parameters.
  \end{minipage}
\end{table}

The panel encompasses 110 LDA--delivery-year cells with observed clearing
prices across 17 LDAs. The three benchmark calibration years used throughout
Sections~\ref{sec:model} and~\ref{sec:results} are selected to represent
three distinct capacity market regimes: \textit{2023/24}, a low-price year
under the old VRR design with no evidence of strategic withholding
(\$34.13 per MW-day, $\mathrm{RSI}_3 = 0.68$); \textit{2025/26}, a
high-price year under the old VRR design that preceded the redesign
(\$269.92 per MW-day, $\mathrm{RSI}_3 = 0.62$); and \textit{2026/27},
the first auction under the new VRR design, which cleared at exactly the
Point~(a) cap (\$329.17 per MW-day, $\mathrm{RSI}_3 = 0.64$). Together,
these three years span the full range of price outcomes in the sample and
allow the comparative static analysis to isolate the effect of VRR design
changes from variation in market structure and cost parameters.

% ------------------------------------------------------------------------------
\subsection{Market Structure}
\label{subsec:cal_market_structure}
% ------------------------------------------------------------------------------

Capacity market concentration in PJM is high by any standard measure. The PJM
IMM reports LDA-level Herfindahl-Hirschman Index (HHI) values in its annual
State of the Market (SotM) reports; in every delivery year covered by our
sample, the RTO-level HHI exceeds 2{,}500 and most constrained LDAs exceed
4{,}000, thresholds the Department of Justice classifies as
highly concentrated markets under standard antitrust benchmarks.

The key calibration input is the number of strategic sellers, $K$. We use the
Residual Supply Index with threshold three (RSI$_3$) as the identifying
criterion: RSI$_3 < 1$ means that the three largest sellers collectively hold
pivotal capacity, i.e., no competitive outcome is supportable without their
participation. In all three benchmark years, the RTO-level RSI$_3$ falls below
unity (0.68, 0.62, and 0.64 for 2023/24, 2025/26, and 2026/27, respectively),
establishing that at most three large sellers are jointly pivotal. The
2026/27 at-cap clearing outcome further rules out $K \geq 4$: our model
predicts interior clearing for four or more strategic sellers in that year
(Section~\ref{subsec:res_concentration}), so the observed cap clearing
is only consistent with $K \leq 3$. The IMM TPS test confirms individual
pivotality for all three largest sellers in every constrained LDA, supporting
$K = 3$ over $K = 2$ as the baseline. We therefore set $K = 3$ as the
baseline strategic count, and examine sensitivity to this assumption via
the $K$ comparative static in Section~\ref{subsec:res_concentration}. Because RSI data is available only at the RTO level
for most delivery years, we impose the same $K = 3$ structure on each LDA
calibration and allow cross-LDA variation to enter through the VRR parameters
and CETL-based fringe capacity.\footnote{LDA-level TPS test outcomes reported
in the IMM SotM confirm that at least three suppliers pass the pivotal test in
all constrained LDAs, consistent with $K = 3$.}

% ------------------------------------------------------------------------------
\subsection{Cost Parameters}
\label{subsec:cal_cost}
% ------------------------------------------------------------------------------

The avoidable cost rate (ACR) for each strategic seller is set to
$c = \$150$ per MW-day, the combined-cycle sector average reported in the 2025
PJM IMM State of the Market (Table 3-8). Combined-cycle units represent the
marginal capacity technology in PJM's current resource mix, and the IMM
identifies the combined-cycle ACR as the relevant benchmark for capacity-offer
mitigation under the TPS test. The symmetric-firm assumption sets all $K = 3$
sellers to the same $c$; heterogeneity in costs and capacity sizes is
abstracted away. It is important to note that the ACR is a proxy for
opportunity cost, not its direct measure. The true avoidable cost of
offering a megawatt of capacity depends on projected energy-market revenues,
fuel hedge positions, maintenance schedules, and the firm's financial
structure --- none of which are publicly observable at the portfolio level.
The sector-average combined-cycle ACR therefore understates cost
heterogeneity across strategic sellers. The sensitivity of equilibrium
markups to this approximation is examined in
Section~\ref{subsec:res_transmission}, where fringe supply is scaled by a
factor $\gamma \in [0.5, 2.0]$ around the baseline; this implicitly tests
robustness to a wide range of effective cost assumptions.

Fringe supply is modeled as an inelastic block offered at $c$: any capacity
not controlled by the $K$ strategic sellers is assumed to offer at cost.
At the RTO level, fringe capacity is set to total installed capacity minus the
capacity attributed to the three largest sellers; this residual is scaled with
$\gamma$ to explore fringe-size sensitivity. For individual LDA calibrations,
the Capacity Emergency Transfer Limit (CETL) serves as the effective fringe
capacity, since CETL measures the maximum deliverable import from the rest of
PJM into the LDA. LDA-level CETL values by delivery year are drawn directly
from the PJM planning parameter files described in
Section~\ref{subsec:cal_data}.

% ------------------------------------------------------------------------------
\subsection{Benchmark Delivery Years}
\label{subsec:cal_years}
% ------------------------------------------------------------------------------

We select three delivery years as calibration benchmarks, each representing
a distinct capacity market regime. \textit{2023/24} is a low-price interior
year (\$34.13/MW-day) under the old VRR design in which PJM's TPS mitigation
was broadly applied; the gap between the model's unmitigated SFE price and the
observed clearing price provides a clean counterfactual for the cost of
mitigation inaction. \textit{2025/26} is a high-price year (\$269.92/MW-day)
that preceded the VRR redesign; it exhibits a large upward price spike relative
to prior years and partial TPS application, making it useful for isolating the
effect of VRR slope on equilibrium markups. \textit{2026/27} is the first
auction under the new VRR design and cleared exactly at the Point~(a) cap
(\$329.17/MW-day), providing the natural experiment used in
Section~\ref{subsec:res_baseline} to identify $K = 3$ as the effective
strategic count.

We exclude \textit{2024/25} from the benchmark set. PJM conducted the 2024/25
BRA as a spot auction held only 0.8 months before the delivery year began,
compared to the 34--36 months of lead time typical of other auctions
(Table~\ref{tab:bra_timeline}). This compressed timeline eliminates any
forward investment signal and makes the 2024/25 clearing price (\$28.92/MW-day)
an unreliable guide to equilibrium capacity incentives; the SFE framework, which
embeds a long-run investment signal in the VRR demand curve, is not applicable.
