% ==============================================================================
% Section: Cap-Incidence and the Emergence of Shortage
% Status: REFRAMED (2026-05-02) — laser-focus rewrite. Shortage-emergence
%   framing made explicit; bunching paragraph absorbed; diagnostic subsection
%   added; cross-refs updated to new structure.
% ==============================================================================

\section{Cap-Incidence and the Emergence of Shortage}
\label{sec:cap_incidence}

This section documents which administrative constraint was binding in
each of PJM's seven completed Base Residual Auctions and shows that a
quantity shortfall against the reliability requirement emerged precisely
when the Shapiro settlement cap began to bind. The textbook prediction
of Section~\ref{subsec:theory_ceiling}---that a binding ceiling on a
tight market produces a quantity shortfall against the planning
target---is matched by the timing of the observed shortfall in PJM.

\input{tables/tab_cap_incidence}

Three regimes are visible in Table~\ref{tab:cap_incidence}, distinguished
by the joint pattern of clearing price, reserve excess relative to the
Installed Reserve Margin (IRM), and IMM-applied mitigation activity.

The first regime, covering 2021/22 through 2024/25, is one of comfortable
supply margins and cost-anchored clearing. Reserve excess IRM was
strongly positive in 2023/24 and 2024/25 (+5{,}687~MW and +3{,}278~MW;
data unavailable for 2021/22 and 2022/23). Clearing prices ranged from
\$29 to \$140/MW-day, well below both the sector-average Net ACR and
the design-level VRR Point~(a) cap of \$391--\$482/MW-day. TPS
mitigation was applied to 0--7\% of resources, but with the marginal
cleared offer well below any active cap, mitigation operative for a
minority of supra-marginal resources did not pin the clearing outcome.

The second regime, in 2025/26, is a transition. Reserve excess IRM
turned slightly negative for the first time (\textminus205~MW), the
clearing price jumped from \$29 to \$270/MW-day, and TPS mitigation
rose to 5.5\% of resources. Yet \$270 remains far below the design
Point~(a) of \$452, so no administrative cap was binding the clearing.
2025/26 is the auction in which the underlying supply--demand balance
tightened materially without any cap yet binding the headline outcome,
and---importantly---the auction still procured adequate capacity.

The third regime, comprising 2026/27 and 2027/28, is one of cap-binding
clearing under negative reserve excess. Both auctions cleared at the
new-design VRR Point~(a) value, which under the Shapiro/PJM settlement
embodies the agreed maximum price. Reserve excess IRM was
\textminus492~MW in 2026/27 and \textminus6{,}517~MW in 2027/28; the
IMM declared both auctions ``not competitive''
\citep{MonitoringAnalytics2025_sotm}. The drop in TPS mitigation
application from 6.3\% in 2026/27 to 1.0\% in 2027/28 is consistent with
the settlement cap moving sufficiently close to the underlying offer
distribution that fewer individual offers exceeded their unit-specific
Net ACR caps. The shortfall relative to the reliability requirement
\emph{emerged in the same years that the cap became binding}, not before.

The IMM-published distribution of resources across offer-cap categories
is consistent with the cap-incidence reading. Across the sample,
Capacity Performance offers fall predominantly into two clusters:
price-takers offering at \$0 (12--35\% of resources) and resources at
their cost-basis cap (57--81\%), together accounting for 88--93\% of
resources in every year. This bimodality is a structural feature of the
mitigation rules, not a behavioral choice: the Default ACR offer cap is
a hard ceiling for any resource subject to mitigation, and the
price-taker option is the only alternative for resources unwilling to
compute and defend a unit-specific cap. The settlement cap, when
binding, truncates the upper cluster: a resource whose cost-basis cap
exceeds \$325 cannot clear above \$325 even if its unit-specific
calculated cap would otherwise allow it. Whether the settlement induced
\emph{additional} clustering near \$325/MW-day cannot be determined from
the aggregated IMM categories; unit-level offer histograms will not be
public until 2029.

The cap-incidence panel offers a direct answer to a recurring question
about the post-settlement auctions: did seller behavior change between
the slack and tight regimes, or did the binding administrative
constraint change while the underlying competitive process remained
similar? The data favor the latter interpretation. In slack years,
clearings of \$29--\$140/MW-day fell well below any plausible
unconstrained strategic price (cost-anchored offers and TPS
mitigation pinned the headline price). In the cap-binding years,
clearing equaled the cap by construction in any model in which the
at-cap regime obtains. The dramatic price changes between regimes
therefore reflect, at least in part, a change in which administrative
constraint binds, not a structural shift in conduct. The independent
evidence that this is not a circular reading is the \emph{timing} of
the reserve-margin flip: reserve excess IRM turned negative in the
2025/26 transition auction---one year before the cap took
effect---and stayed negative in the two cap-binding years that
followed.

This reading does not contradict the IMM's substantive assessment. The
IMM attributes the underlying scarcity in 2026/27 and 2027/28 to data
center load growth rather than to any change in seller behavior or to
the settlement itself \citep{MonitoringAnalytics2025_sotm}. The
contribution of this section is to make that interpretation visible at
a glance and to frame the dollar consequences---developed for the two
capped auctions in Section~\ref{sec:case_studies}---as the result of
overlaying a binding administrative cap on a market entering scarcity,
rather than as the result of strategic conduct that the cap then
disciplined.
