% ==============================================================================
% Section: Conclusion
% Status: COMPLETE (2026-03-28 Session R6 rewrite)
% ==============================================================================

\section{Conclusion}
\label{sec:conclusion}

On December 30, 2024, Pennsylvania filed a FERC complaint arguing that PJM's
capacity auction demand curve was miscalibrated given near-term entry
expectations. One month later, Governor Shapiro announced a settlement: the
VRR curve for the 2026/27 and 2027/28 auctions would be reshaped, with a cap
at \$325/MW-day UCAP and a floor at \$175/MW-day UCAP, at a projected savings
of over \$21~billion for 65~million consumers. This paper uses a calibrated
Supply Function Equilibrium model to evaluate the settlement's terms, its
stated counterfactual, and its consequences for entry incentives and
information.

Five findings emerge.

\textbf{First, the settlement is binding under the observed market structure.}
With $K \leq 3$ pivotal suppliers (RSI$_3 < 1$ in every benchmark year
is consistent with an upper bound of three, and the TPS test showing three
individually pivotal sellers supports $K = 3$ over $K = 2$), the SFE equilibrium for 2026/27 falls at the VRR
Point~(a) cap of \$329.17/MW-day, exceeding the settlement cap
$\bar{p}_S = \$325$ by \$4.17/MW-day at the RTO level. The at-cap outcome
is consistent with $K \leq 3$: four or more symmetric strategic sellers would
predict interior clearing at \$228 or below (Section~\ref{subsec:res_concentration}).
In constrained LDAs, where the 2025/26 BRA produced prices of
\$444--\$466/MW-day, the structural gap between the SFE equilibrium and the
settlement cap is considerably larger. The settlement reduces the price
paid in the 2026/27 and 2027/28 auctions. It does not reduce the price-cost
margin $L$, which remains 0.54 under the strategic equilibrium.

\textbf{Second, the settlement modifies the demand curve, not merely the
price outcome.} The settlement does not impose an external overlay. It
replaces the VRR curve shape with a flat top at \$325, the existing sloped
segment between \$325 and \$175, and a flat bottom at \$175 that procures
all sub-threshold supply at the floor price. The \$175 floor is an
investment-adequacy mechanism (below the RTO Net CONE of \$212/MW-day in
2026/27) designed to prevent the price from collapsing in over-supplied
future auctions. Applying a floor while capping the price creates an
administered price band that is neither the competitive equilibrium
(\$150/MW-day) nor the strategic equilibrium (\$329--\$333/MW-day). The
incentive implications of the band depend on where the unconstrained
equilibrium falls relative to the Net CONE threshold, which varies across
LDAs (Section~\ref{subsec:alloc_floor}).

\textbf{Third, the \$21~billion savings figure uses the wrong counterfactual.}
Shapiro's claim is computed as the gap between the unconstrained VRR
Point~(a) price (``over \$500/MW-day'') and the settlement cap, multiplied
by cleared quantity over two delivery years. The SFE model provides a
structural alternative: the strategic equilibrium $p^*$ against the
unmodified VRR curve. Using RTO-level quantities, the VRR counterfactual
implies a two-year transfer of approximately \$17.16~billion at the
RTO level (constrained-LDA effects account for the gap to the \$21~billion
headline, though the precise LDA contribution requires LDA-level cleared
quantity data not analyzed here). The SFE
counterfactual implies a two-year transfer of approximately \$612~million
at the RTO level (Section~\ref{sec:21billion}). The two estimates answer different questions.
The VRR-based figure measures the gap between the design ceiling and the
settlement cap, a policy decision about how to calibrate the demand curve.
The SFE-based figure measures the revenue difference between the model's
uncapped equilibrium benchmark and the settlement cap under the maintained
assumptions. The first is 35~times larger than the second because the
2026/27 strategic equilibrium happens to fall near the design ceiling, not
because the settlement eliminates large strategic rents.

\textbf{Fourth, the lead-time argument for the settlement is partial.}
The complaint's investment-adequacy theory, that the VRR cap is too high
given near-term entry expectations, is strongest for greenfield generation
(36--48~month deployment) and weakest for demand response, storage, and
existing-plant upgrades (6--18~months). For the 2026/27 BRA held 10.2~months
ahead, at least some capacity categories could have responded to price
signals. More fundamentally, prices inform investment even when direct
construction is infeasible within the lead time: through retirement deterrence,
mobile resource deployment, portfolio contracting, regulatory signaling, and
the formation of expectations for future delivery years
(Section~\ref{subsec:lt_channels}). The correct structural remedy for
lead-time compression is schedule restoration, which PJM is pursuing, not
price suppression.

\textbf{Fifth, structural alternatives dominate price controls.}
Three interventions reduce SFE equilibrium prices without suppressing the
scarcity signal: CETL expansion (30\% fringe increase $\to$ $-$24~pp
price-cost margin reduction; 10~pp CETL/reliability-requirement increase $\to$
4--6~pp reduction); VRR slope steepening beyond the ER25-1357 baseline
(slope factor $\geq 1.4$ moves 2026/27 from cap-binding to interior clearing
even at $K = 3$); and entry promotion to reach $K \geq 4$ (which alone
would reduce 2026/27 prices by \$101/MW-day without any demand-curve
modification). Each of these options is permanent and preserves the price
signal, whereas the settlement is temporary and suppresses it
(Section~\ref{sec:policy_alternatives}).

The central finding is that PJM's 2026/27 and 2027/28 capacity prices are
high in a manner consistent with structurally thin supply and tight
transmission constraints: three sellers hold pivotal capacity in every
benchmark year, and CETL limits restrict competitive imports in the most
constrained LDAs. The settlement
addresses the symptom (high prices) rather than the cause (concentrated
supply with limited competitive fringe). It is defensible as emergency relief
during a period of acute auction-schedule disruption. It is not a substitute
for the transmission investment, VRR redesign, and entry promotion that would
make a similar intervention unnecessary in future auctions.

Several directions for future work follow naturally from the model's
limitations. An asymmetric-firm SFE would characterize how cost and capacity
heterogeneity interact with the TPS screen and the settlement's LDA application.
A dynamic model with endogenous entry and exit would connect the short-run
price-cost margin findings to long-run capacity adequacy under the price band.
An empirical structural model recovering firm-level cost parameters from
observed offer data would allow direct welfare calculation and would permit
testing whether the $K = 3$ symmetric calibration understates or overstates
actual strategic markups. The calibrated SFE results in this paper provide a
structural benchmark for each of these extensions and a quantitative foundation
for evaluating the settlement's long-run consequences as the 2028/29 auction
cycle approaches.
