% ==============================================================================
% Section: Discussion
% Status: COMPLETE (2026-03-28)
% ==============================================================================

\section{Discussion}
\label{sec:discussion}

% ------------------------------------------------------------------------------
\subsection{Comparison to IMM Metrics}
\label{subsec:disc_imm}
% ------------------------------------------------------------------------------

The PJM IMM relies primarily on the Three Pivotal Supplier (TPS) test and
associated Residual Supply Index to assess market power in each BRA. The TPS
test is a binary pass/fail screen: if RSI$_3 < 1$, the three largest sellers
are jointly pivotal and their offers are capped at the ACR level. This design
is well-suited for detecting the \textit{presence} of market power and applying
offer-price mitigation in real time, but it provides no quantitative estimate
of the \textit{magnitude} of the markup that strategic sellers would impose in
the absence of mitigation.

Our calibrated SFE framework fills this gap. For all three benchmark years, the
model predicts Lerner indices of 0.54--0.65 at $K = 3$. Translating to a
price-cost ratio, this implies equilibrium prices 2.2--2.9 times the
competitive level. The 2026/27 at-cap result is the most informative for identification: the
Holmberg boundary condition mechanically selects $p^* = \bar{p}$ when $K = 3$
and capacity constraints bind, confirming that the 2026/27 market structure
is consistent with $K = 3$ strategic sellers at the capacity ceiling. As
discussed in Section~\ref{subsec:res_baseline}, the informativeness is in the
at-cap versus interior binary, not the dollar amount of the clearing price. The IMM's TPS test
correctly identifies all three years as requiring mitigation (RSI$_3 < 1$ in
each case), but the binary outcome understates the severity of the market power
problem: a Lerner index of 0.60 corresponds to a markup of \$90--\$180/MW-day above
marginal cost depending on the year --- a magnitude that HHI concentration
indices, being structural rather than behavioral measures, cannot directly
quantify.

The LDA-level results reinforce this picture. All 41 LDA-year cells in our
sample yield Lerner indices above 0.48, with the median near 0.60. The TPS
test results reported in Table~\ref{tab:tps_results} show consistently
constrained LDAs, but the heterogeneity in Lerner values across LDAs ---
driven by import penetration (CETL/reliability requirement) --- is not
captured by any existing IMM metric. Our cross-LDA regression in
Section~\ref{subsec:res_lda} finds that each ten-percentage-point increase in
import penetration reduces the Lerner index by approximately five percentage
points, a structural relationship that is invisible to the TPS screen.

% ------------------------------------------------------------------------------
\subsection{Policy Implications}
\label{subsec:disc_policy}
% ------------------------------------------------------------------------------

Three policy levers follow directly from the analysis.

\paragraph{VRR Curve Design.}
The comparative static in Section~\ref{subsec:res_vrr} shows that
steepening the VRR demand curve (increasing $|D'(p)|$) reduces equilibrium
markups by raising the elasticity of demand facing each seller. Doubling the
VRR slope from the old-design baseline reduces the Lerner index by 20--30
percentage points in interior-clearing years. FERC Order ER25-1357, which
mandated a transition from the three-point to the four-point VRR design,
moves in this direction for EMAAC and several LDAs. However, the new design
also introduces an explicit floor price of approximately \$177--\$179/MW-day
at Points~(b) and~(c). The 2026/27 calibration illustrates the binding nature
of this floor: the equilibrium price is determined not by the ODE interior
solution but by the Point~(a) cap, precisely because the floor prevents the
downward-sloping portion of the demand curve from accommodating competitive
entry at low prices. Steepening the demand curve while simultaneously raising
the price floor may therefore produce limited markup reduction for high-power
LDAs.

\paragraph{Transmission Expansion and CETL.}
The cross-LDA evidence in Section~\ref{subsec:res_lda} points to import
capacity --- as measured by CETL --- as the most important structural
determinant of market power across LDAs. DEOK, with import penetration near
unity, approaches competitive outcomes despite having the same $K = 3$
strategic structure as other LDAs. Expanding CETL through transmission
investment in constrained zones (PS, EMAAC, COMED) would reduce effective
market concentration without requiring any change to auction rules. Our
point estimate implies that a 10-percentage-point increase in import
penetration reduces the Lerner index by roughly five percentage points; the
effect is likely nonlinear at high penetration levels but directionally
robust. Transmission policy is therefore a capacity-market policy.

\paragraph{Structural Remedies for High-Power LDAs.}
Several LDAs --- including PS, EMAAC, and COMED --- exhibit Lerner indices
above 0.60 across all benchmark years and are structurally incapable of
achieving competitive outcomes with three strategic sellers and limited
imports. For these LDAs, offer-price mitigation based on ACR caps provides
incomplete protection because sellers retain the option to withhold
\textit{quantity} rather than price. A more robust remedy would combine TPS
mitigation with minimum-offer quantity requirements or structural divestiture
to reduce $K$ to fewer than three in the most constrained zones. Our model provides
a quantitative benchmark: the Lerner index at the observed $K$ gives the
markup that mitigation must eliminate to restore competitive outcomes.

% ------------------------------------------------------------------------------
\subsection{Limitations}
\label{subsec:disc_limitations}
% ------------------------------------------------------------------------------

Several modeling choices bound the scope of our conclusions. First, we impose
symmetric firms, as discussed and defended in Section~\ref{subsec:model_symmetric}.
The reported Lerner values should be read as capacity-weighted average markups;
actual sellers with below-average costs exercise more market power, and those
with above-average costs exercise less. The asymmetric extension requires
numerical solution of $K$ coupled ODEs \citep{Anderson2008_sfe_asymmetric},
which we leave for future work.

Second, our model is static: each auction is solved independently, and there
is no dynamic entry, exit, or contracting. This limitation is more than
technical. PJM's 2026/27 price spike coincides with a rapid acceleration in
data-center load growth across the PJM footprint. If structural demand growth
was outpacing capacity additions, the high clearing price could partly reflect
a legitimate dynamic scarcity signal --- the kind of investment incentive
that capacity markets are explicitly designed to provide --- rather than pure
rent extraction by strategic sellers. The static SFE model attributes the
entire gap between $p^*$ and $c$ to markup; it cannot separate the strategic
component from the scarcity component. Importantly, the model's $K = 3$
identification does not resolve this ambiguity: the at-cap structural test
and the RSI$_3$ criterion identify pivotal market power regardless of whether
the scarcity is demand-driven or supply-driven. Decomposing strategic markups
from dynamic scarcity rents would require a long-run SFE with endogenous
investment \citep{Joskow2007_capacity}, which we leave for future work.

Third, we abstract from demand-side uncertainty. The VRR curve is treated as
deterministic, whereas actual PJM auctions involve uncertainty over reliability
requirements and BRA participation. Incorporating stochastic demand would
smooth the kink in the VRR schedule and likely reduce the quantitative
difference between old- and new-design Lerner estimates.

Finally, the ACR of \$150/MW-day is a technology average for combined-cycle
units. Firm-specific ACRs can deviate substantially from this benchmark,
particularly for peaking units (which face higher ACRs) or nuclear resources
(which are not subject to ACR-based mitigation). The fringe-supply sensitivity
analysis in Section~\ref{subsec:res_transmission} provides a partial check on
ACR uncertainty, showing that Lerner values are robust to a factor-of-two
variation in fringe capacity size.
