% ==============================================================================
% Section: Institutional Background
% Status: COMPLETE (2026-03-27)
% ==============================================================================

\section{Institutional Background}
\label{sec:institutional}

% ------------------------------------------------------------------------------
\subsection{PJM's Reliability Pricing Model}
\label{subsec:inst_rpm}
% ------------------------------------------------------------------------------

PJM Interconnection operates the largest competitive wholesale electricity
market in North America, coordinating the dispatch of more than 180{,}000
megawatts of generating capacity across thirteen states and the District of
Columbia. Unlike energy markets, which settle the hourly exchange of
electricity, capacity markets procure commitments from generators to be
available during high-demand periods in a future delivery year. These
commitments address the ``missing money'' problem: energy market price caps
suppress the scarcity rents that would otherwise finance investment in peaking
generation, creating a structural shortfall in incentives to build new
capacity \citep{Joskow2007_capacity, Joskow2008_capacity_payments}.

PJM implemented the Reliability Pricing Model (RPM) in 2007 to provide a
forward capacity market with explicit locational pricing
\citep{Bowring2013_pjm}. The primary procurement mechanism is the Base
Residual Auction (BRA), in which suppliers submit capacity offers and the
market clears against a downward-sloping administrative demand schedule---the
Variable Resource Requirement (VRR) curve---at a uniform clearing
price.\footnote{Capacity Performance reforms introduced in 2015--2016 created
a two-product structure (Base Capacity and Capacity Performance), but the BRA
mechanism and the VRR curve remain central to price formation throughout our
sample.} PJM also conducts Incremental Auctions in the months preceding each
delivery year to accommodate late capacity additions and withdrawals, but
these represent only a small fraction of total procurement.

Capacity procurement occurs separately for Locational Deliverability Areas
(LDAs), geographic sub-regions of the PJM footprint constrained by
transmission import limits. Each LDA has its own VRR curve parameterized by
the LDA's local Cost of New Entry (Net CONE) and reliability requirement, and
clears at a price at or above the RTO-wide clearing price. The transmission
capacity available to import from outside an LDA is governed by the Capacity
Emergency Transfer Limit (CETL). Our panel covers the RTO and sixteen
sub-LDAs across delivery years 2021/22 through 2027/28, producing 110
LDA--delivery-year observations with observed clearing prices.

% ------------------------------------------------------------------------------
\subsection{The Variable Resource Requirement Curve}
\label{subsec:inst_vrr}
% ------------------------------------------------------------------------------

The VRR curve is the administrative demand schedule that PJM presents to
sellers in each BRA. Its shape is determined by two parameters: the Net Cost
of New Entry (Net CONE) and the reliability requirement. Net CONE represents
the annualized cost of building and operating a reference generating unit,
net of projected revenues from energy and ancillary service markets. It is
computed independently for each LDA and updated each delivery year based on
cost surveys and projected market revenues.

Under the old VRR design in place for delivery years 2021/22 through
2025/26, the schedule was defined by three anchor points.\footnote{We
distinguish the ``old'' design from the ``new'' design introduced by FERC
Order ER25-1357 for delivery years 2026/27 and beyond; see
Section~\ref{subsec:inst_break}.} Point~(a), the price cap, was set at
$1.5 \times \text{Net CONE}$ at a quantity slightly below the reliability
requirement. Point~(b) was set at $0.75 \times \text{Net CONE}$ at the
reliability requirement itself. Point~(c) set the floor price to zero at a
quantity slightly above the reliability requirement. Between anchor points the
schedule is interpolated linearly, producing a steeply downward-sloping
segment from Point~(a) to Point~(b) and a shallower segment from Point~(b)
to the zero-price floor.

The economic rationale for a downward-sloping demand schedule, rather than a
vertical demand set at the reliability requirement, is to moderate the extreme
price volatility that vertical demand generates when supply is inelastic near
the reliability threshold \citep{Cramton2005_capacity, Hobbs2007_rpm}. A
sloped demand schedule trades some quantity uncertainty for price stability:
the clearing price falls smoothly as surplus capacity rises, rather than
collapsing discontinuously from the cap to zero once the reliability
requirement is met. This design was specifically advocated as a capacity
market reform to improve investment incentives and dampen boom-bust price
cycles \citep{CramtonOckenfels2012_capacity}.

Because the VRR parameters are publicly announced well in advance of the
auction, the demand side of the market is fully transparent to sellers. Every
participant observes the price-quantity pairs of all anchor points before
submitting offers. This transparency is a defining feature of the PJM
capacity market and distinguishes it from environments with demand
uncertainty---a distinction that proves consequential for equilibrium
selection in the SFE framework developed in Section~\ref{sec:model}.

% ------------------------------------------------------------------------------
\subsection{The 2026 VRR Redesign}
\label{subsec:inst_break}
% ------------------------------------------------------------------------------

FERC approved a restructured VRR curve design in 2025 (Order ER25-1357),
taking effect with the 2026/27 delivery year BRA held in July 2025. The
redesign introduced a fourth anchor point that creates an explicit price
floor. Under the new design:

\begin{itemize}
  \item Point~(a) remains the price cap, set at approximately $1.5 \times
    \text{Net CONE}$;
  \item A downward-sloping segment connects Point~(a) to Point~(b), the
    floor price (approximately \$177--\$179 per MW-day in our sample);
  \item A flat segment at the floor price extends from Point~(b) to
    Point~(c), so that additional capacity clears at the floor rather than
    at zero;
  \item Point~(d) marks the end of the flat floor segment at a quantity well
    above the reliability requirement.
\end{itemize}

The key economic change is the elimination of the collapse to a zero price
floor. By setting a hard lower bound of approximately \$177--\$179 per
MW-day---close to the projected net energy and ancillary service revenue
available to a new entrant---the redesign prevents prices from falling below
the level at which new investment would break even, even in conditions of
surplus capacity.

Table~\ref{tab:bra_timeline} documents the BRA timeline and RTO clearing
outcomes for all seven completed auctions in our sample. Two facts stand out.
First, both auctions conducted under the new VRR design---the 2026/27 BRA
(July 2025) and the 2027/28 BRA (January 2026)---cleared at exactly the
Point~(a) price cap: \$329.17 and \$333.44 per MW-day, respectively. This
``at-cap'' outcome, in which aggregate supply intersects the VRR at its
uppermost anchor point, is consistent with strategic equilibrium pricing
and motivates the supply function equilibrium analysis in
Sections~\ref{sec:model}--\ref{sec:results}. Second, the RTO clearing price
rose sharply between 2024/25 (\$28.92 per MW-day) and 2025/26 (\$269.92),
a price spike that precedes the VRR redesign and suggests a structural shift
in capacity market conditions beginning with the 2025/26 delivery year. The
contrast between the old-design auctions---which exhibited outcomes ranging
from \$28.92 to \$269.92 per MW-day---and the new-design auctions, which
uniformly cleared at the cap, provides a natural experiment for the
comparative static in Section~\ref{subsec:res_vrr}.

\begin{table}[H]
  \centering
  \caption{PJM Base Residual Auction Timeline and RTO Clearing Outcomes,
    2021/22--2027/28}
  \label{tab:bra_timeline}
  \begin{tabular}{lcccccl}
    \toprule
    Delivery & BRA    & Lead   & Clearing    & Net CONE    & VRR    & At  \\
    Year     & Date   & (mo.)  & (\$/MW-day) & (\$/MW-day) & Design & Cap \\
    \midrule
    2021/22 & May 2018    & 37.0 & \$140.00 & \$321.57 & Old & No \\
    2022/23 & May 2021    & 12.5 & \$50.00  & \$260.50 & Old & No \\
    2023/24 & July 2022   & 11.0 & \$34.13  & \$274.96 & Old & No \\
    2024/25 & May 2024    &  0.8 & \$28.92  & \$293.19 & Old & No \\
    2025/26 & August 2024 &  9.9 & \$269.92 & \$228.81 & Old & No \\
    2026/27 & July 2025   & 10.2 & \$329.17 & \$212.14 & New & \textbf{Yes} \\
    2027/28 & January 2026 & 16.8 & \$333.44 & \$242.52 & New & \textbf{Yes} \\
    \bottomrule
  \end{tabular}
  \vspace{3pt}

  \begin{minipage}{0.92\textwidth}
    \footnotesize \textit{Note:} Lead time is the number of calendar months
    between the BRA settlement date and the start of the delivery year. RTO
    clearing prices and Net CONE values are in nominal dollars per MW-day.
    ``At Cap'' indicates that the RTO clearing price equaled the Point~(a)
    price cap under the relevant VRR design. Data sources: PJM Planning
    Parameters; PJM BRA Results.
  \end{minipage}
\end{table}

% ------------------------------------------------------------------------------
\subsection{Auction Lead-Time Irregularity}
\label{subsec:inst_leadtime}
% ------------------------------------------------------------------------------

A second distinctive feature of the PJM capacity market during this period
is the extreme irregularity of the BRA schedule. In a well-functioning
capacity market, auctions should be held well in advance of the delivery
year---typically two to three years ahead---to provide investment signals
with enough lead time for new generation to be financed, permitted, and
constructed \citep{Hobbs2007_rpm}. The PJM RPM was designed with a three-year
forward horizon.

Table~\ref{tab:bra_timeline} reveals substantial deviations from this design.
The 2021/22 BRA was held in May 2018, 37 months before the start of
delivery---consistent with the intended horizon. However, a series of
FERC-mandated market reforms delayed subsequent auctions, compressing lead
times to 12.5 months for 2022/23 and 11 months for 2023/24. No BRA was held
at all in calendar year 2023. Most strikingly, the 2024/25 BRA was conducted
in May 2024, a mere 0.8 months before the June 2024 start of delivery---
effectively a spot capacity auction. Under such conditions, no new generation
could plausibly have been built in response to the price signal; the auction
served purely as a spot-market procurement exercise rather than an investment
coordination mechanism.

The disrupted schedule has two implications for our analysis. First, the
2024/25 delivery year should not be interpreted as a period in which the
capacity market was providing forward investment incentives, and the low
clearing price of \$28.92 per MW-day that year reflects the absence of a
forward capacity shortage rather than an equilibrium outcome of a properly
functioning forward market. Second, the compressed lead times for 2022/23
through 2024/25 may have contributed to the sharp price increase observed in
2025/26: if new entrants were unable to respond to scarcity signals in time
for those delivery years, the accumulated capacity shortfall would manifest
as high prices in subsequent auctions. Our SFE simulation takes the VRR
curve and capacity endowments as given and does not model the investment
response to price signals, but this context is important when interpreting
the 2024/25 and 2025/26 results.

% ------------------------------------------------------------------------------
\subsection{Market Power Mitigation: The Three Pivotal Supplier Test}
\label{subsec:inst_tps}
% ------------------------------------------------------------------------------

PJM's primary market power mitigation mechanism in capacity auctions is the
Three Pivotal Supplier (TPS) test. Supplier $i$ is defined as
\emph{pivotal} in LDA $\ell$ if removing $i$'s capacity from the market
would cause aggregate supply to fall short of the reliability requirement---
that is, if the market cannot clear without $i$'s participation. The TPS
test asks whether any individual seller, or any coalition of the three largest
sellers, is pivotal. When the test is triggered, the affected suppliers'
offers are capped at the Avoidable Cost Rate (ACR), a measure of short-run
variable cost \citep{Bowring2013_pjm}.

The Residual Supply Index for $k$ suppliers in LDA $\ell$ is defined as
\begin{equation}
  \mathrm{RSI}_{k}^{\ell} \;=\;
  \frac{\text{Total capacity in LDA }\ell
        \;-\; \text{Capacity of top-}k\text{ sellers in LDA }\ell}
       {\text{Reliability requirement for LDA }\ell}.
  \label{eq:rsi}
\end{equation}
An RSI below unity means the market outside the top-$k$ sellers cannot meet
the reliability requirement on its own, confirming that those sellers are
collectively pivotal. Mitigation is triggered whenever $\mathrm{RSI}_1 < 1$
or $\mathrm{RSI}_3 < 1$.

Table~\ref{tab:tps_results} summarizes TPS test outcomes for the RTO and
selected LDAs across the delivery years for which PJM IMM State of the Market
reports provide RSI data. Three patterns stand out. First, $\mathrm{RSI}_3$
falls below unity for every LDA and every delivery year in the panel, so
the TPS mitigation threshold is met throughout our entire sample. Second, the
concentration is strikingly persistent: in the RTO, $\mathrm{RSI}_3$ ranges
only from 0.62 to 0.73 across six delivery years despite substantial
year-to-year variation in clearing prices (\$28.92 to \$333.44 per MW-day).
Third, the most extreme concentration appears in small sub-LDAs: BGE and
DEOK each report a single seller with $\mathrm{RSI}_1 = 0$ throughout the
sample, implying that those LDAs depend entirely on one supplier to meet
their reliability requirement.

\begin{table}[H]
  \centering
  \caption{TPS Test Results: Residual Supply Indices by LDA and Delivery Year}
  \label{tab:tps_results}
  \begin{tabular}{llcccc}
    \toprule
    Delivery Year & LDA   & $\mathrm{RSI}_1$ & $\mathrm{RSI}_3$ &
      Participants & All Pivotal \\
    \midrule
    2022/23 & RTO   & 0.81 & 0.73 & 130 & Yes \\
    2022/23 & MAAC  & 0.69 & 0.37 &  25 & Yes \\
    2022/23 & EMAAC & 1.25 & 0.64 &   7 & Yes \\
    2022/23 & COMED & 0.43 & 0.36 &  14 & Yes \\
    2022/23 & BGE   & 0.00 & 0.00 &   1 & Yes \\
    2022/23 & DEOK  & 0.00 & 0.00 &   1 & Yes \\
    \midrule
    2023/24 & RTO   & 0.78 & 0.68 & 134 & Yes \\
    2023/24 & MAAC  & 0.78 & 0.40 &  11 & Yes \\
    \midrule
    2024/25 & RTO   & 0.77 & 0.64 & 133 & Yes \\
    2024/25 & MAAC  & 0.59 & 0.11 &   9 & Yes \\
    2024/25 & EMAAC & 0.48 & 0.00 &   2 & Yes \\
    \midrule
    2025/26 & RTO   & 0.82 & 0.62 & 128 & Yes \\
    2026/27 & RTO   & 0.82 & 0.64 & 153 & Yes \\
    2027/28 & RTO   & 0.82 & 0.63 & 155 & Yes \\
    \bottomrule
  \end{tabular}
  \vspace{3pt}

  \begin{minipage}{0.92\textwidth}
    \footnotesize \textit{Note:} $\mathrm{RSI}_k$ is the Residual Supply
    Index for the top-$k$ sellers; see equation~\eqref{eq:rsi}. ``All
    Pivotal'' indicates that every participant is individually pivotal
    (i.e., $\mathrm{RSI}_1 < 1$ when computed for each seller individually).
    Data source: PJM IMM State of the Market reports, various years.
  \end{minipage}
\end{table}

Critically, the TPS test is a binary instrument: it identifies whether
mitigation conditions are met but says nothing about the magnitude of
strategic markup or the size of the resulting price-cost margin. In
the RTO, $\mathrm{RSI}_3$ has been persistently low and essentially stable
across six delivery years---yet clearing prices have ranged from \$28.92 to
\$333.44 per MW-day, a factor of eleven. The binary TPS test cannot
distinguish the near-competitive outcome in 2023/24 from the at-cap outcomes
in 2026/27 and 2027/28. The supply function equilibrium framework developed
in Section~\ref{sec:model} provides a structural model that generates
continuous, quantitative price-cost margin estimates and can explain this
price variation through changes in the VRR curve parameters and the market
structure documented here.
