% ==============================================================================
% Section: Introduction
% Status: COMPLETE (2026-03-28 Session 3 rewrite)
% ==============================================================================

\section{Introduction}
\label{sec:introduction}

On December 30, 2024, the Commonwealth of Pennsylvania filed a complaint
with the Federal Energy Regulatory Commission alleging that PJM's capacity
auction demand curve was miscalibrated: its price cap was set too high given
near-term expectations for new entry. The complaint cited PJM's July 2024
Base Residual Auction for the 2025/26 delivery year, which cleared at
\$269.92/MW-day --- an increase of more than 800 percent from the prior year
--- as evidence that the existing market design imposed unnecessary costs on
65 million consumers across 13 states \citep{Shapiro2025_press_release}.
Six weeks later, on January 28, 2025, Governor Josh Shapiro announced an
agreement with PJM: the demand curves for the 2026/27 and 2027/28 auctions
would be modified to impose a price cap of approximately \$325/MW-day
(unforced capacity, UCAP) and a price floor of approximately \$175/MW-day
UCAP, applied to both the RTO and all constrained Locational Deliverability
Areas \citep{PJM2025_settlement_slides}. The agreement was projected to save
consumers over \$21 billion over the two delivery years.

The settlement creates a two-tier price ceiling in PJM's capacity market.
PJM's existing Variable Resource Requirement (VRR) curve already contains a
design-level cap at Point~(a), set at approximately 1.5 times the Net Cost
of New Entry --- an investment-adequacy ceiling calibrated to the annualized
cost of a reference generating unit and intended to prevent prices from
permanently exceeding the entry threshold. The settlement imposes a second,
lower cap at $\bar{p}_S \approx \$325$/MW-day, below the Point~(a) level,
while simultaneously adding a floor at $\bar{p}_f \approx \$175$/MW-day
that clears all capacity offered at or below that level. The mechanism is
implemented by reshaping the demand curve itself: a flat top at
$\bar{p}_S$, the existing sloped segment between $\bar{p}_S$ and
$\bar{p}_f$, and a flat bottom at $\bar{p}_f$ \citep{PJM2025_settlement_slides}.
This paper asks: what price would PJM's 2026/27 and 2027/28 auctions have
cleared at absent the settlement modification, and what are the consequences
of suppressing prices to $\bar{p}_S$?

We answer the question using a calibrated Supply Function Equilibrium (SFE)
model of the PJM Base Residual Auction. In the model, $K$ symmetric
strategic sellers simultaneously submit supply functions against PJM's VRR
demand curve, which is publicly known before bidding. Following
\citet{Klemperer1989_sfe} and \citet{Green1992_british}, equilibrium is
characterized by a first-order ordinary differential equation (ODE) system
in the sellers' supply functions, integrated backward from the
capacity-constrained boundary. Equilibrium selection follows a Holmberg-motivated boundary condition:
at the VRR Point~(a) cap, each seller's supply function equals its full
installed capacity. This criterion is motivated by \citet{Holmberg2008_unique_sfe}'s
uniqueness result for stochastic markets; in the deterministic BRA it is
imposed as a selection device, and the at-cap clearing outcomes in 2026/27
and 2027/28 confirm the condition is active in practice. We calibrate the model to three
benchmark delivery years --- 2023/24 (low price, old VRR), 2025/26 (price
spike, old VRR), and 2026/27 (at-cap, new VRR) --- using published PJM
planning parameters, BRA clearing results, and IMM State of the Market cost
data. The SFE equilibrium price $p^*$ is the counterfactual: where the
unmodified VRR curve would have cleared under strategic bidding. This
counterfactual is what is needed to evaluate the settlement's claim of
\$21 billion in savings.

Four findings emerge. First, under the observed market structure ---
$K = 3$ pivotal suppliers with residual supply indices below unity ---
SFE equilibrium prices reach the VRR Point~(a) cap in 2026/27 and exceed
the settlement cap $\bar{p}_S = \$325$/MW-day. Price-cost margins (Lerner
indices) are 0.54--0.65 across benchmark years. The at-cap 2026/27 outcome
is consistent with $K \leq 3$: four or more symmetric sellers would predict
interior clearing below the cap, so the observed at-cap outcome is
consistent with a low effective strategic-seller count. Second, the settlement is
binding. The SFE equilibrium $p^* = \$329$/MW-day at the RTO level and
substantially higher in constrained LDAs (where 2025/26 prices reached
\$444--\$466/MW-day); in both cases, equilibrium prices exceed
$\bar{p}_S = \$325$/MW-day. Suppressing prices from $p^*$ to $\bar{p}_S$
distorts entry incentives and destroys information about scarcity conditions
\citep{Hayek1945_knowledge}. Third, the settlement's \$21 billion savings
claim uses the unconstrained VRR Point~(a) price (approximately
\$500/MW-day) as the counterfactual; our SFE model provides a structural
alternative estimate that is lower but still exceeds the settlement cap.
The difference between these counterfactuals is economically consequential:
the Point~(a) counterfactual includes scarcity rents that are part of the
market design, while the SFE model provides a calibrated market-clearing
benchmark under the model's assumptions --- not a clean decomposition of
scarcity versus conduct, since the static model cannot fully separate the two. Fourth, transmission expansion --- modeled as increases
in the Capacity Emergency Transfer Limit (CETL) --- organically reduces
equilibrium prices by shifting LDAs from cap-binding to interior-clearing
regimes, addressing the structural source of high prices without suppressing
price signals \citep{Weitzman1974_prices}.

The remainder of the paper is organized as follows.
Section~\ref{sec:literature} surveys the supply function equilibrium
literature, the capacity market design literature, and the economics of
price controls in oligopolistic markets. Section~\ref{sec:institutional}
reviews PJM's institutional context, including the VRR mechanism, auction
irregularities, and the Shapiro settlement. Section~\ref{sec:model}
develops the theoretical SFE model. Section~\ref{sec:calibration} describes
data sources and calibration. Section~\ref{sec:results} presents SFE
equilibrium prices and comparative statics relative to the settlement cap.
Section~\ref{sec:allocative_cost} analyzes the allocative consequences of
the settlement's price band for entry, exit, and information.
Section~\ref{sec:leadtime} evaluates the lead-time argument underlying
the complaint. Section~\ref{sec:21billion} examines the \$21~billion
savings claim. Section~\ref{sec:policy_alternatives} considers structural
alternatives to price caps. Section~\ref{sec:conclusion} concludes.
