% ==============================================================================
% Section: Literature Review
% Paper: IO Paper 2 -- Calibrated SFE Simulation of Market Power in PJM RPM
% Status: SKELETON -- headings and \cite{} only, no prose
% Last updated: 2026-03-24
% ==============================================================================

\section{Literature Review}
\label{sec:literature}

% ------------------------------------------------------------------------------
\subsection{Supply Function Equilibria: Theory}
\label{subsec:lit_sfe_theory}
% ------------------------------------------------------------------------------

The supply function equilibrium concept, introduced by
\citet{Klemperer1989_sfe}, considers an oligopoly in which each firm commits
to a schedule mapping prices to quantities before demand is realized. Unlike
Bertrand competition (commitment to a fixed price) or Cournot competition
(commitment to a fixed quantity), the supply function strategy allows each
firm to adapt its output to whatever market price emerges after the demand
shock is realized. \citeauthor{Klemperer1989_sfe} characterize the symmetric
Nash equilibria of this game under additive demand uncertainty, showing that
equilibria are characterized by a first-order differential equation whose
solutions form a continuum: in general, any supply function tracing a
trajectory between the competitive schedule (price equals marginal cost) and
the Cournot schedule (as if rivals bid at capacity) constitutes a Nash
equilibrium. The specific equilibrium realized depends on market coordination,
and demand uncertainty is what disciplines firms to choose upward-sloping
supply functions over fixed prices.

\citet{Green1992_british} apply the Klemperer-Meyer framework to the England
and Wales electricity spot market after privatization, modeling two dominant
generators as strategic suppliers to a pool settled at a uniform price.
Their calibrated duopoly shows that the Nash equilibrium in supply functions
implies substantial markups above marginal cost---with prices near the top of
the feasible equilibrium range---and that entry of additional firms (or
divestiture into five units) would significantly reduce those markups. Green
and Newbery establish the methodological template followed here: express
demand as a known function of price, derive the symmetric first-order ODE,
and integrate backward from the capacity-constrained boundary to obtain the
equilibrium supply function. Their paper is the closest methodological
antecedent to the present work, with the key difference that our demand
function is PJM's explicit piecewise-linear VRR curve rather than a load-
duration curve, and our market clears annually in a single forward auction
rather than half-hourly in a spot pool.

The multiplicity problem endemic to SFE is resolved by
\citet{Holmberg2008_unique_sfe}, who shows that under conditions natural to
capacity-constrained electricity markets---symmetric producers, inelastic
demand, a price cap, and capacity constraints that bind with positive
probability---the supply function equilibrium is unique. The key insight is
that when rivals' capacity constraints are simultaneously binding, firm $i$
faces perfectly inelastic residual demand; the monopoly optimum at that
moment is to charge exactly the price cap. This ties the boundary condition
of the ODE: each firm's supply function must equal its full installed
capacity at the price cap. The arbitrary constant in the ODE solution is
thereby pinned, selecting a unique trajectory. This result is central to the
identification strategy of the present paper, since the VRR Point~(a) price
serves as the price cap and the at-cap clearing outcomes in 2026/27 and
2027/28 are precisely the event Holmberg requires to be positive probability.
\citet{Holmberg2009_paybid} extends the uniqueness analysis to pay-as-bid
auctions, showing that a unique symmetric equilibrium obtains there as well.
\citet{Vasin2016_sfe_uniform} confirm existence and characterize additional
properties of symmetric SFE in uniform-price settings with oligopolistic
supply.

The effect of forward commitments on equilibrium supply functions is
examined by \citet{Allaz1993_forward}, who show in a Cournot setting that
forward markets reduce equilibrium markups through a Stackelberg-style
commitment effect. \citet{Anderson2005_sfe_pricecaps} extend the SFE framework to settings
with both forward contracts and price caps, establishing existence conditions
for equilibria under these combined constraints. Capacity commitments
in the BRA are formally analogous to forward sales, so the static SFE applied
to the auction game incorporates these disciplining effects into the
equilibrium concept.

% ------------------------------------------------------------------------------
\subsection{Numerical Methods for SFE}
\label{subsec:lit_sfe_methods}
% ------------------------------------------------------------------------------

The numerical computation of SFE relies on the differential equation
characterization of equilibrium supply functions. For the symmetric case the
ODE reduces to a single equation integrable backward from the capacity
boundary. For asymmetric firms the system of $K$ coupled ODEs requires
iterative methods. \citet{Anderson2008_sfe_asymmetric} develop the leading
algorithm for the asymmetric case: initialize all firms at competitive supply
functions, solve each firm's single-firm ODE taking rivals' current functions
as fixed, update simultaneously, and iterate to convergence. They establish
that equilibrium solutions can be ordered---a partial ordering that guides
the search for the strong equilibrium in which all but one firm are at
capacity at peak demand---and demonstrate convergence using GAMS
implementations for calibrated electricity market examples. The
Anderson--Hu algorithm is the primary computational reference for the
asymmetric extension described in Section~\ref{subsec:cal_market_structure}.

\citet{Baldick2004_linear_sfe} develop the analytically tractable special
case in which both the demand curve and firms' marginal cost functions are
linear. Under these restrictions the symmetric ODE has a closed-form
solution parameterized by a single scalar, yielding explicit comparative
statics with respect to the number of firms and the demand slope. Linear SFE
provides a useful calibration benchmark and validates numerical solvers, but
imposes functional form restrictions violated by the piecewise-linear VRR
curve in the PJM application. \citet{Niu2005_sfe_forward} extend the
numerical framework to generators holding fixed forward contracts, and
\citet{Rudkevich2004_sfe_applications} develop a piecewise-linear
approximation scheme that handles non-differentiabilities at capacity
constraints. In this paper we follow the backward-integration approach,
numerically solving the symmetric ODE using the \texttt{deSolve} package in
R with the Holmberg boundary condition as the starting point.

The empirical performance of SFE simulations is evaluated by
\citet{Sioshansi2007_sfe_ercot}, who compare simulated equilibria to observed
prices in the ERCOT balancing market. They find that SFE tracks observed
prices well at high demand but overpredicts markups at low demand, consistent
with theory: strategic withholding is most profitable when residual demand is
inelastic. This asymmetry also appears in our results, where the at-cap
outcomes under the new VRR design---when the entire VRR curve is above the
floor and residual demand is effectively inelastic---correspond to the most
extreme predicted markups.

% ------------------------------------------------------------------------------
\subsection{Market Power in Electricity Markets}
\label{subsec:lit_electricity_mp}
% ------------------------------------------------------------------------------

The literature on market power in restructured electricity markets has developed
two complementary methodologies: counterfactual simulation and structural
estimation. \citet{Borenstein2002_market_power} establish the counterfactual
simulation template by decomposing the price-cost gap in California's
restructured wholesale market into market power and scarcity components. Using
a competitive benchmark derived from reported marginal costs, they find that
market power accounts for a substantial share of the observed price-cost wedge,
particularly during high-demand hours when residual demand is inelastic.
Their decomposition method --- simulate competitive prices under the observed
cost and demand conditions, then attribute the gap to strategic behavior ---
is the direct antecedent to the approach used in this paper for capacity
auctions. \citet{Wolak2003_measuring} extends this tradition by developing
a unilateral market power measure derived directly from electricity bid
data, exploiting the observation that a pivotal supplier's residual demand
curve is effectively inelastic at the market equilibrium. The pivotal supplier
concept underlying Wolak's measure is the same logic codified in PJM's
Residual Supply Index and Three Pivotal Supplier test.

\citet{Wolfram1999_duopoly} provides the first empirical test of SFE
predictions, using detailed bid data from the British electricity spot market
duopoly (National Power and PowerGen) to compare observed offers against the
Nash equilibrium predictions of the Klemperer-Meyer model. She finds that
actual markups fall between the Cournot and competitive benchmarks, broadly
consistent with the range of SFE equilibria, but closer to Cournot than the
symmetric Nash equilibrium would predict. \citet{Sweeting2007_market_power}
revisits the England and Wales market over a longer horizon (1995--2000) using
a structural model and finds that offer behavior converged toward SFE
predictions as generators accumulated experience with the market mechanism.
Together, Wolfram and Sweeting establish that SFE provides empirically useful
predictions for concentrated electricity spot markets, motivating its
application here to the forward capacity auction setting.

\citet{Bushnell2008_vertical} extend the simulation approach to multiple US
electricity markets and introduce vertical integration --- between generation
and retail supply --- as a structural determinant of effective market
concentration. Vertically integrated firms sell power to their own retail
customers at regulated rates, reducing the incentive to withhold generation
and thereby attenuating spot-market market power. Their cross-market evidence
establishes that the structural features of seller positions (portfolio
composition, contracting, vertical ownership) govern markup magnitudes as
much as the number of firms. \citet{Wolak2007_quantifying} provides
complementary evidence from the Australian NEM that forward contracts
directly discipline spot bids, consistent with the Allaz-Vila mechanism
discussed in Section~\ref{subsec:lit_sfe_theory}. The capacity commitment
in PJM's BRA creates a form of forward obligation that in principle disciplines
the subsequent capacity-auction markup --- though our results suggest this
discipline is incomplete when the committed positions are themselves held by
the same pivotal suppliers.

% ------------------------------------------------------------------------------
\subsection{Residual Demand and the Baker--Bresnahan Framework}
\label{subsec:lit_residual_demand}
% ------------------------------------------------------------------------------

\citet{Baker1988_residual_demand} develop the foundational econometric
framework for estimating the residual demand curve facing a single firm. The
key insight is that a firm's market power is fully characterized by the
elasticity of the residual demand it faces: the Lerner index equals the
negative inverse of that elasticity at the equilibrium quantity. Estimating
the residual demand curve --- total demand minus the supply forthcoming from
rival firms at each price level --- requires instruments that shift rivals'
costs without directly affecting the firm's own cost or demand. Baker and
Bresnahan propose cost shifters for rivals as instruments and apply the
framework to the US brewing industry.

In the capacity auction context, the residual demand curve facing a strategic
seller is a shifted version of PJM's VRR schedule: total VRR demand minus the
aggregate supply of fringe and rival-strategic sellers at each price. The
critical distinguishing feature of the PJM application is that the residual
demand schedule is set \textit{administratively} by the regulator and
published before each auction; sellers observe the VRR parameters directly.
The Baker-Bresnahan problem of recovering the residual demand from market
data does not arise here --- the demand curve is known in closed form. This
observation is what enables the SFE calibration strategy: given the observed
VRR schedule and the number of strategic sellers identified by the RSI$_3$
test, the equilibrium supply function and implied Lerner index can be computed
without instruments or demand estimation.

% ------------------------------------------------------------------------------
\subsection{Capacity Market Design}
\label{subsec:lit_capacity_markets}
% ------------------------------------------------------------------------------

The economics literature on capacity market design originates from a
fundamental coordination problem: competitive energy markets, subject to
regulatory price caps, tend to suppress the scarcity rents that would
otherwise finance investment in peaking generation. \citet{Joskow2007_capacity}
formalize this ``missing money'' problem and demonstrate that even under
competitive conditions, energy-only markets underinvest in reliability when
price caps bind during scarcity events. \citet{Joskow2008_capacity_payments}
examines the remedies, arguing that capacity payments can correct the
underinvestment provided they are designed to replicate missing scarcity rents
without creating perverse entry or exit incentives.

A central design question is whether the procurement target should be fixed
at the reliability requirement---a vertical demand schedule---or
downward-sloping, as in PJM's VRR mechanism.
\citet{Cramton2005_capacity} make the case for a sloped administrative demand
curve on the grounds that a vertical schedule generates extreme price
volatility near the reliability threshold: small fluctuations in supply
produce large swings between the price cap and zero. A sloped demand curve
trades some quantity variation for price stability and provides smoother
investment incentives across the capacity cycle. \citet{Hobbs2007_rpm}
provides a formal dynamic analysis of the PJM Reliability Pricing Model and
confirm that the VRR curve substantially reduces price volatility relative to
vertical demand, at the cost of modest deviations from the target reliability
level.

The uniform-price format of the BRA has also received attention.
\citet{Cramton2007_uniform} defend the uniform-price auction over pay-as-bid
alternatives, arguing that under the former, sellers have weaker incentives
to shade their bids strategically and equilibrium prices are closer to the
competitive outcome. \citet{CramtonOckenfels2012_capacity} extend this
analysis to the broader landscape of capacity market design and examine the
interaction between auction format and market power. These papers identify
a tension at the heart of capacity markets: the sloped administrative demand
curve that dampens boom-bust price cycles is precisely the curve whose shape
governs the rent-extraction opportunities available to a strategic seller.

The market power dimension of capacity auctions is examined in detail by
\citet{Bowring2013_pjm}, who provides an overview of PJM's RPM with attention
to the role of the Independent Market Monitor in detecting and mitigating
anticompetitive behavior. Bowring documents the application of the TPS test
and ACR mitigation across delivery years and LDAs, and notes that market
power concerns are particularly acute in constrained LDAs where import
capacity is limited. \citet{Vossler2009_price_caps} use laboratory experiments
to study behavioral responses to soft price caps in uniform-price auction
environments and find that caps do bind strategically---sellers' offers cluster
near the cap, consistent with a focal equilibrium at the ceiling. Taken
together, this literature motivates a structural approach to measuring the
strategic markup that the binary TPS test can only detect but not quantify.

% ------------------------------------------------------------------------------
\subsection{Price Controls in Oligopolistic Markets}
\label{subsec:lit_price_controls}
% ------------------------------------------------------------------------------

The economics of price controls in oligopolistic markets provides the
theoretical foundation for the allocative analysis in
Section~\ref{sec:allocative_cost}. The foundational reference is
\citet{Weitzman1974_prices}, who shows that the choice between price
instruments and quantity instruments for regulatory intervention depends
on the relative slopes of the marginal benefit and marginal cost schedules.
When the marginal cost schedule is steep (supply is inelastic in the short
run) and the marginal benefit schedule is relatively flat (demand is elastic),
price instruments dominate quantity instruments because they avoid the
deadweight loss from misallocating constrained supply. Capacity markets
present precisely this structure: short-run supply is inelastic (capacity
is fixed within the delivery year) and the administrative demand curve has
a downward-sloping segment whose elasticity the regulator can calibrate.
Weitzman's analysis implies that modifying the VRR price ceiling is an
efficient regulatory instrument if and only if the unconstrained equilibrium
price is genuinely above the socially optimal level --- a question our SFE
model addresses directly by providing a structural estimate of $p^*$.

\citet{GlaeserLuttmer2003_rent} demonstrate that binding price controls in
markets with imperfect competition generate misallocation that is
qualitatively different from the standard deadweight loss triangle. When
quantity must be rationed below the unconstrained equilibrium, goods are
allocated to those with the strongest relationships to sellers rather than
to those with the highest willingness to pay, creating efficiency losses
that are a multiple of the simple triangle. In capacity markets the
analogue is resource allocation across time: a binding cap suppresses the
scarcity signal that should direct investment toward future delivery years
with highest expected returns, creating an efficiency loss that accumulates
dynamically rather than appearing as a static deadweight triangle.

In the SFE literature, \citet{Anderson2005_sfe_pricecaps} extend the
Klemperer-Meyer framework to incorporate both price caps and forward
contracts, establishing conditions under which equilibria with caps exist
and characterizing how caps modify the ODE boundary condition. Their
analysis shows that a price cap does not simply truncate the equilibrium
supply function: it changes the strategic problem by altering the payoff at
the capacity boundary. When the cap is below the unconstrained Holmberg
boundary price, firms face a modified optimization at the cap level, and the
equilibrium supply function differs from the uncapped case throughout the
price range, not just at the ceiling. This result is directly relevant to
the PJM application: the settlement does not merely cap revenues at
\$325/MW-day, it modifies the entire strategic incentive structure facing
the $K = 3$ sellers.

\citet{Vossler2009_price_caps} provide experimental evidence that price caps
induce focal behavior in uniform-price auction environments: offers cluster
near the cap even when the unconstrained equilibrium would clear below it.
Applied to PJM, this suggests that the settlement floor at \$175/MW-day
may itself become a focal point in future auctions with ample supply, as
sellers coordinate on offering just above the floor rather than competing
down to marginal cost. Finally, \citet{Cramton2005_capacity} emphasize that
the missing money problem --- the chronic revenue shortfall that motivates
capacity markets in the first place --- is recreated if capacity market
prices are suppressed below the investment threshold. The settlement floor
at \$175/MW-day represents a partial recognition of this risk: PJM
characterizes the floor as ``supporting near-term investment,'' an implicit
acknowledgment that the cap side of the settlement poses the very problem
that the capacity market was designed to solve.
