% ==============================================================================
% Section: Model
% Status: COMPLETE (2026-03-27)
% ==============================================================================

\section{Model}
\label{sec:model}

% ------------------------------------------------------------------------------
\subsection{Setup}
\label{subsec:model_setup}
% ------------------------------------------------------------------------------

The BRA is modeled as a static, complete-information game in which sellers
simultaneously submit supply functions before the auction clears. This
formulation follows \citet{Klemperer1989_sfe} and, for the electricity market
application, \citet{Green1992_british}. The timing is: (1) PJM publishes the
VRR curve parameters for the upcoming delivery year; (2) sellers submit
supply functions; (3) PJM clears the auction at the uniform price where
aggregate supply meets the VRR curve; (4) sellers deliver committed capacity
during the delivery year.

\begin{assumption}
  \label{assm:players}
  There are $K \geq 2$ strategic sellers, indexed $i = 1, \ldots, K$, and a
  competitive fringe. Each strategic seller $i$ holds a fixed capacity
  portfolio $\bar{q}_i > 0$ measured in megawatts. A supply function
  $S_i : [0, \bar{p}] \to [0, \bar{q}_i]$ is a nondecreasing, twice
  continuously differentiable mapping from price to quantity. Each seller
  simultaneously commits to a supply function before the auction clears.
\end{assumption}

\begin{assumption}
  \label{assm:fringe}
  The competitive fringe bids at its avoidable cost rate $c_f > 0$. Fringe
  supply is $S_f(p) = 0$ for $p < c_f$ and $S_f(p) = \bar{Q}_f$ for
  $p \geq c_f$, where $\bar{Q}_f > 0$ is total fringe capacity. The fringe
  is price-taking and its behavior is common knowledge.
\end{assumption}

\begin{assumption}
  \label{assm:cost}
  Each strategic seller has a constant marginal cost $c > 0$ for output
  $q \leq \bar{q}_i$ and infinite marginal cost above capacity. For the
  symmetric calibration we set $c$ equal to the sector-average avoidable
  cost rate published by the PJM IMM.
\end{assumption}

The market clears at the price $p^*$ satisfying
\begin{equation}
  \sum_{i=1}^{K} S_i(p^*) + S_f(p^*) = D(p^*),
  \label{eq:clearing}
\end{equation}
where $D(\cdot)$ is the VRR demand schedule described below. Every seller
receives the same clearing price $p^*$ for its committed capacity. The
\emph{residual demand} facing seller $i$ at price $p$ is
\begin{equation}
  R_i(p) \;=\; D(p) - \sum_{j \neq i} S_j(p) - S_f(p).
  \label{eq:resid}
\end{equation}
In equilibrium, $S_i(p^*) = R_i(p^*)$. Because $D(\cdot)$ and all rivals'
strategies are publicly known before the auction, there is no demand
uncertainty: the residual demand curve each seller faces is deterministic.%
\footnote{Holmberg's \citeyear{Holmberg2008_unique_sfe} uniqueness proof
  formally requires stochastic demand with positive probability of all
  capacity being demanded. In the deterministic BRA setting, the boundary
  condition $s(\bar{p}) = \bar{q}$ is imposed as a selection criterion
  motivated by that result. The at-cap clearing outcomes in 2026/27 and
  2027/28 confirm the condition is active in practice.}
This determinism means firms observe the VRR parameters before submitting
offers; the boundary condition is imposed as an equilibrium selection
criterion, as discussed in Section~\ref{subsec:model_selection}.

% ------------------------------------------------------------------------------
\subsection{The VRR Demand Schedule}
\label{subsec:model_vrr}
% ------------------------------------------------------------------------------

PJM's Variable Resource Requirement curve is a piecewise-linear,
publicly-announced downward-sloping demand schedule. Under the old design
(delivery years 2021/22--2025/26) it is parameterized by three anchor points
$(p_a, q_a)$, $(p_b, q_b)$, and $(0, q_c)$:
\begin{equation}
  D(p) =
  \begin{cases}
    q_a                                              & p \geq p_a \\[4pt]
    q_a + \dfrac{q_b - q_a}{p_b - p_a}(p - p_a)   & p_b \leq p < p_a \\[6pt]
    q_b + \dfrac{q_c - q_b}{0 - p_b}(p - p_b)     & 0 \leq p < p_b,
  \end{cases}
  \label{eq:vrr_old}
\end{equation}
where $p_a = 1.5 \times \text{Net CONE}$ is the price cap and
$p_b = 0.75 \times \text{Net CONE}$ is the reliability-requirement price.
The quantities $q_a < q_b < q_c$ are the corresponding capacity levels, with
$q_b$ equal to the reliability requirement. Because $D(p) = q_a$ for all
$p \geq p_a$, the VRR curve is effectively inelastic above the cap.

Under the new design (delivery years 2026/27 onward), FERC Order ER25-1357
introduces a fourth anchor point $(p_f, q_f)$ and $(p_f, q_d)$ with
$q_f > q_b$, creating a flat floor segment at price $p_f \approx \$177$--
$\$179$/MW-day. The new schedule is:
\begin{equation}
  D(p) =
  \begin{cases}
    q_a                                              & p \geq p_a \\[4pt]
    q_a + \dfrac{q_b - q_a}{p_f - p_a}(p - p_a)   & p_f \leq p < p_a \\[6pt]
    q_d                                              & 0 \leq p < p_f.
  \end{cases}
  \label{eq:vrr_new}
\end{equation}
The flat floor segment at $p_f$ means that clearing prices cannot fall below
$p_f$ along the VRR, changing the elasticity profile that sellers face and
thus the equilibrium markup. The comparative static in
Section~\ref{subsec:res_vrr} isolates this design change.

The slope $D'(p)$ is piecewise constant and satisfies:

\begin{assumption}
  \label{assm:demand}
  $D'(p) < 0$ on the sloped segments; $D'(p) = 0$ on the flat segments
  (at and above $p_a$, and below $p_f$ under the new design). The function
  $D(\cdot)$ is known to all players before the auction.
\end{assumption}

% ------------------------------------------------------------------------------
\subsection{Firm Optimization and the SFE Condition}
\label{subsec:model_foc}
% ------------------------------------------------------------------------------

Seller $i$ chooses $S_i(\cdot)$ to maximize profit. Taking rivals'
supply functions as fixed, the profit at any market-clearing price $p$ is
$\pi_i = p \cdot S_i(p) - c \cdot S_i(p) = (p - c) \cdot S_i(p)$, since
marginal cost is constant at $c$. The seller's problem is equivalent to
maximizing profit along its residual demand curve $R_i(\cdot)$. At
price $p$, if seller $i$ considers increasing its offered quantity by
$\mathrm{d}q$, the market-clearing price must fall by $\mathrm{d}p$ to
preserve equation~\eqref{eq:clearing}:
\begin{equation}
  \mathrm{d}p
  \;=\; \frac{\mathrm{d}q}{D'(p) - \sum_{j \neq i} S_j'(p) - S_f'(p)}.
  \label{eq:dprice}
\end{equation}
Since $D'(p) < 0$ and $S_j'(p) \geq 0$, the denominator is strictly
negative: an increase in supply lowers the clearing price. The seller's
marginal revenue from increasing quantity at price $p$ is:
\begin{equation}
  \text{MR}_i(p)
  \;=\; p + \frac{S_i(p)}{D'(p) - \sum_{j \neq i} S_j'(p) - S_f'(p)}.
  \label{eq:mr}
\end{equation}
Setting marginal revenue equal to marginal cost $c$ yields the
\emph{supply function equilibrium condition}:

\begin{proposition}
  \label{prop:sfe_foc}
  In a supply function equilibrium, each seller $i$'s strategy
  $S_i(\cdot)$ satisfies
  \begin{equation}
    S_i(p) \;=\; -\bigl(p - c\bigr)
    \Bigl[D'(p) - \textstyle\sum_{j \neq i} S_j'(p) - S_f'(p)\Bigr]
    \label{eq:sfe_foc}
  \end{equation}
  at every price $p$ at which the market can clear. This is a system of
  $K$ coupled first-order ordinary differential equations in the supply
  functions $\{S_1, \ldots, S_K\}$.
\end{proposition}

Equation~\eqref{eq:sfe_foc} is the $K$-seller generalization of the
first-order condition derived by \citet{Green1992_british} for the symmetric
duopoly (their equation~3) and of the general condition in
\citet{Anderson2008_sfe_asymmetric} (their equation~1). Its economic content
is straightforward: seller $i$ equates its residual-demand markup $(p - c)$
to the inverse of the residual demand elasticity at that price. When more
sellers compete ($K$ large) or the VRR slope is steep (large $|D'|$), the
system requires a smaller markup at each price, pushing the equilibrium
supply function toward the competitive schedule.

% ------------------------------------------------------------------------------
\subsection{Equilibrium Selection and the Boundary Condition}
\label{subsec:model_selection}
% ------------------------------------------------------------------------------

The ODE system~\eqref{eq:sfe_foc} generally admits a continuum of solutions.
\citet{Klemperer1989_sfe} show that under demand uncertainty, the set of
symmetric supply function equilibria is an interval bounded above by the
Cournot schedule (where each firm behaves as if rivals are at capacity) and
below by the competitive schedule (where $p = c$). Any trajectory between
these two stationary solutions is a Nash equilibrium. This multiplicity
complicates both prediction and comparative statics.

The multiplicity is resolved by the following result, which applies
directly to the capacity-auction setting:

\begin{proposition}[\citealt{Holmberg2008_unique_sfe}]
  \label{prop:holmberg}
  Suppose sellers are symmetric, demand is inelastic above the price cap
  $\bar{p}$, and the probability that aggregate demand reaches the capacity
  ceiling is strictly positive. Then there exists a unique symmetric supply
  function equilibrium. In this equilibrium each seller's supply function
  satisfies the boundary condition
  \begin{equation}
    S_i(\bar{p}) \;=\; \bar{q}_i \quad \text{for all } i.
    \label{eq:bc}
  \end{equation}
\end{proposition}

The intuition is direct. When rivals' capacity constraints simultaneously
bind, seller $i$ faces perfectly inelastic residual demand. Its monopoly
optimum at that moment is to set price equal to the cap $\bar{p}$; any lower
price sacrifices revenue without reducing quantity (since residual demand is
inelastic), and any higher price is infeasible. This forces the supply
function to pass through the point $(\bar{p},\, \bar{q}_i)$, tying the
arbitrary constant in the ODE and selecting a unique trajectory.

In the PJM BRA, the price cap $\bar{p}$ equals the VRR Point~(a) price, and
Assumption~\ref{assm:demand} ensures $D'(\bar{p}) = 0$ (the VRR is flat at
the cap). The Holmberg conditions are met: PJM's reliability target requires
essentially all installed capacity, so there is positive probability that
aggregate supply reaches the ceiling at $\bar{p}$. The at-cap clearing
outcomes in 2026/27 and 2027/28 (Table~\ref{tab:bra_timeline}) confirm that
the boundary condition is active in those years. We therefore integrate the
ODE~\eqref{eq:sfe_foc} \emph{backward} from the boundary condition
$S_i(\bar{p}) = \bar{q}_i$, following the solution method of
\citet{Green1992_british}.

% ------------------------------------------------------------------------------
\subsection{The Symmetric Special Case}
\label{subsec:model_symmetric}
% ------------------------------------------------------------------------------

For calibration and comparative statics we begin with the symmetric case in
which all $K$ strategic sellers are identical: $\bar{q}_i = \bar{q}$ and
$c_i = c$ for all $i$. In a symmetric Nash equilibrium,
$S_i(p) = s(p)$ for all $i$. Substituting into
equation~\eqref{eq:sfe_foc}:
\begin{equation}
  s(p) \;=\; -\bigl(p - c\bigr)
    \Bigl[D'(p) - (K-1)\,s'(p) - S_f'(p)\Bigr].
  \label{eq:sfe_sym_raw}
\end{equation}
Under Assumption~\ref{assm:fringe}, $S_f'(p) = 0$ almost everywhere.
Rearranging~\eqref{eq:sfe_sym_raw} for $s'(p)$ yields the \emph{symmetric
SFE ordinary differential equation}:
\begin{equation}
  \boxed{
    s'(p) \;=\; \frac{1}{K-1}
    \left[ D'(p) + \frac{s(p)}{p - c} \right],
    \quad p \in [c,\, \bar{p}),
  }
  \label{eq:sfe_ode}
\end{equation}
with boundary condition $s(\bar{p}) = \bar{q}$. This is a first-order,
nonautonomous ODE in the single unknown function $s(\cdot)$. It is
the $K$-firm generalization of \citeauthor{Green1992_british}'s
equation~(3), to which it reduces when $K = 2$ and $S_f \equiv 0$.

Equation~\eqref{eq:sfe_ode} has a direct economic interpretation.
The term $D'(p)$ captures the direct effect of price on quantity demanded:
a steeper VRR slope (larger $|D'|$) makes $s'(p)$ more negative, meaning
each firm's equilibrium supply function falls more steeply toward the cap as
price rises. The term $s(p)/(p-c)$ is a markup-boundary feedback: as $p$
approaches $c$ from above, this term diverges and forces $s'(p)$ to remain
bounded away from the demand slope, enforcing the lower boundary of the
supply function and preventing equilibrium offers below cost. At prices well
above $c$, the feedback term partially offsets $D'(p)$, making the supply
function steeper in price than a linear interpolation between $c$ and
$\bar{p}$ would predict. More competitors (larger $K$) divide the
denominator, flattening each firm's supply function toward the competitive
benchmark.

The equilibrium clearing price $p^*$ is found by solving
\begin{equation}
  K \cdot s(p^*) + S_f(p^*) = D(p^*)
  \label{eq:clearing_sym}
\end{equation}
for $p^*$ after integrating~\eqref{eq:sfe_ode} backward from $\bar{p}$.
The implied price-cost margin (Lerner index) is
\begin{equation}
  L \;=\; \frac{p^* - c}{p^*},
  \label{eq:lerner}
\end{equation}
which serves as the primary measure of strategic markup (price-cost margin)
in the results of Section~\ref{sec:results}. The competitive benchmark corresponds to $c = p^*$
($L = 0$), attained as $K \to \infty$ or as $|D'(p)| \to \infty$ (perfectly
elastic demand). The Cournot upper bound on the markup is
$L^{\text{Cournot}} = 1 / (K \cdot |D'(p^*) \cdot p^* / D(p^*)|)$, the
standard inverse-elasticity rule evaluated at market demand (which equals the
residual demand elasticity in the symmetric case).

\paragraph{The symmetry assumption.}
In reality, the three largest PJM capacity sellers hold heterogeneous
portfolios: a mix of baseload nuclear, peaking gas, renewables, and
demand response with widely differing cost structures and capacity sizes.
Treating them as symmetric is a deliberate first-order approximation, not
a statement about the actual market. The justification is tractability: the
symmetric case reduces the $K$-dimensional system of coupled ODEs in
equation~\eqref{eq:sfe_foc} to the single scalar ODE~\eqref{eq:sfe_ode},
making the model calibratable from publicly available aggregate data. This
is the standard approach in the calibrated SFE literature
\citep{Green1992_british, Sioshansi2007_sfe_ercot}. The direction of the
asymmetry bias on the average price-cost margin is ambiguous: a lower-cost
seller has a higher markup than the symmetric benchmark; a
higher-cost seller has a lower one. On average, these effects partially
offset. We treat the symmetric Lerner as an estimate of the
capacity-weighted average markup across the strategic sellers, and note
that the spread around this average is an object of interest for future
asymmetric extensions.

Equation~\eqref{eq:sfe_ode} is solved numerically using a fourth-order
Runge--Kutta (RK4) integrator with fixed step size $h = 0.05$\,\$/MW-day.
To improve stability we apply the substitution $\tau = \bar{p} - p$, which
converts the backward initial-value problem into a standard forward IVP
with $s(0) = \bar{q}$. The piecewise nature of $D'(p)$ is accommodated by
restarting the integration at each VRR kink point: the integration grid is
split at $p_b$ (old design) or $p_f$ (new design), so the integrator never
crosses a discontinuity in $D'(p)$ within a single step. Under the new
design, $D(p)$ has a jump discontinuity at $p_f$ (from $q_b$ approaching
from above to $q_d > q_b$ below); equilibrium is resolved by identifying
$p^*$ as the price at which aggregate supply meets the sloped-segment demand
from above, which coincides with $p_f$ when $K \cdot s(p_f) + S_f(p_f) =
q_b$. Full details of the numerical implementation are provided in
Section~\ref{sec:calibration}.
