% ==============================================================================
% Section: Evaluating the Lead-Time Argument
% Status: COMPLETE (2026-03-28 Session R4)
% ==============================================================================

\section{Evaluating the Lead-Time Argument}
\label{sec:leadtime}

% ------------------------------------------------------------------------------
\subsection{The Complaint's Investment-Adequacy Theory}
\label{subsec:lt_theory}
% ------------------------------------------------------------------------------

The legal theory underlying the Shapiro complaint is not a market-conduct
claim. Pennsylvania did not allege that capacity sellers withheld supply
strategically or bid above competitive levels. The complaint alleged instead
that PJM's VRR curve was ``miscalibrated'' --- that its price cap was
``too high given lower expectations for new entry in the near-term''
\citep{PJM2025_settlement_slides}. The implicit causal chain is: if new
generation cannot be financed, permitted, and constructed before the
delivery year begins, then high capacity prices cannot stimulate entry
within the relevant horizon, and elevated prices serve only as a transfer
from buyers to sellers rather than as an investment signal. On this view,
the demand curve should reflect what consumers are willing to pay for
\emph{available} capacity, not what would be needed to attract capacity
that cannot arrive in time.

This is an investment-adequacy argument about auction design, not a finding
of strategic conduct. It is worth separating the two because the SFE model
speaks directly to the second and only indirectly to the first. The model
confirms that the 2026/27 at-cap outcome is consistent with strategic pricing
by $K \leq 3$ sellers; it does not address whether new entry could have
responded to those prices within the available lead time. The lead-time
argument is therefore independent of the SFE findings --- it is a claim about
the supply side's medium-run elasticity, not about the current-period strategic
equilibrium.

% ------------------------------------------------------------------------------
\subsection{PJM's Lead-Time Record and the Settlement Horizon}
\label{subsec:lt_record}
% ------------------------------------------------------------------------------

Section~\ref{subsec:inst_leadtime} documented the extreme compression of BRA
lead times from 2021/22 through 2024/25. The 2024/25 BRA was held in May
2024 --- just 0.8~months before the June 2024 start of the delivery year ---
making it effectively a spot procurement with no meaningful entry window.
This compression is the empirical foundation for the lead-time argument:
high prices in a spot auction cannot plausibly attract new entry, so the
price cap's calibration to long-run entry costs was disconnected from actual
market conditions.

However, the settlement applies to the 2026/27 and 2027/28 delivery years,
for which the auction schedule had been substantially restored. The 2026/27
BRA was held in July 2025, approximately 10.2~months before the June 2026
delivery start; the 2027/28 BRA was held in January 2026,
approximately 16.8~months ahead \citep{PJM2025_settlement_slides}. These
lead times remain short relative to the 36--48~months required to finance
and construct a new combined-cycle gas plant, but they substantially exceed
the 0.8-month spot horizon that motivates the complaint's sharpest form.

For demand response, battery storage, and capacity upgrades to existing
plants --- resource types with deployment horizons of 6--18~months --- a
10.2-month lead time is non-trivial. The empirical question is whether the
2026/27 price spike (\$269.92/MW-day in the 2025/26 BRA; the 2026/27 BRA
cleared at \$329.17) triggered supply-side responses in these faster-moving
categories before the auction. PJM's Capacity Interconnection Queue and
Demand Response registration data would be needed to answer this directly,
which is beyond the scope of the current analysis. The point is that the
lead-time argument is strongest for greenfield generation and weaker for
incremental and demand-side resources, and these distinctions matter for
evaluating whether the complaint's causal claim holds quantitatively.

% ------------------------------------------------------------------------------
\subsection{Five Channels: Why Prices Still Convey Information}
\label{subsec:lt_channels}
% ------------------------------------------------------------------------------

Even if new greenfield entry is infeasible within a 10-month window, auction
prices serve at least five informational and allocative functions beyond
directly stimulating construction:

\begin{enumerate}

\item \textbf{Retirement deterrence.} High prices reduce the probability
that marginal capacity retires before or during the delivery period.
At \$329/MW-day, a 100~MW generator earns approximately \$12~million per
year from capacity revenue alone --- well above the threshold that would
typically trigger a retirement decision. Capping the price at \$325/MW-day
preserves this incentive nearly intact (the cap is only \$4/MW-day binding
at the RTO level), but the precedent of administrative price suppression
increases uncertainty about future clearing prices, potentially accelerating
retirement decisions in subsequent years.

\item \textbf{Mobile resource deployment.} Demand response programs, behind-
the-meter storage, and contracted interruptible loads can register for
capacity auctions on timescales of weeks to months. High prices provide the
financial incentive to activate these registrations. The settlement cap
reduces the annual payment for a 1~MW DR registration from \$120,085
(at $p^* = \$329$/MW-day) to \$118,625 (at $\bar{p}_S = \$325$/MW-day) ---
a difference of \$1,460/MW-year that is small in absolute terms. The more
consequential signal degradation is forward-looking: if investors anticipate
that future auctions will similarly be capped, the present value of the
registration's upside is truncated for resources evaluating participation
in the 2028/29 auction and beyond.

\item \textbf{Portfolio contracting and hedging.} Load-serving entities and
generation owners enter bilateral capacity contracts in advance of BRAs.
Observed BRA prices inform the forward strip that underlies these contracts.
A cap that suppresses BRA prices therefore also suppresses the contract
prices that hedge investment risk, increasing the cost of capital for
prospective developers regardless of lead time.

\item \textbf{Regulatory and interconnection signaling.} Sustained high
capacity prices signal to state regulators, transmission planners, and
interconnection queue managers that the region is capacity-constrained.
This signal prompts regulatory approvals, permitting, and transmission
investments that improve the supply environment for future auctions. Capping
the price shortens the signal's duration and reduces its magnitude.

\item \textbf{Expectations for future periods.} Investors planning capacity
for 2028/29 and beyond observe 2026/27 and 2027/28 BRA outcomes to form
expectations about future price levels. If the settlement binds in both years,
the observed prices (\$325) understate the strategic equilibrium (\$329--\$333),
leading to under-investment relative to a world where prices reflected
unconstrained scarcity. \citet{Cramton2005_capacity} emphasize that the
capacity market's primary function is to provide a long-run investment signal;
a two-year cap disrupts this function precisely in the years when the signal
is most informative.

\end{enumerate}

% ------------------------------------------------------------------------------
\subsection{The Hayek Critique and the Correct Remedy}
\label{subsec:lt_hayek}
% ------------------------------------------------------------------------------

\citet{Hayek1945_knowledge} argues that the central epistemological problem
of economic organization is the utilization of knowledge that is not given to
any single mind in its totality. The price system solves this problem by
aggregating dispersed information --- about costs, preferences, and scarcity
--- into a summary statistic that enables decentralized coordination. Price
controls suppress this statistic regardless of whether the underlying
information would otherwise have triggered an immediate supply response.
A price cap at \$325/MW-day destroys the signal that $p^* = \$329$ even if
no generator can break ground on a new plant within 10 months.

The appropriate policy response to the lead-time problem is to restore auction
timing, not to suppress auction prices. \citet{Weitzman1974_prices} analyzes
the choice between price and quantity regulatory instruments under uncertainty,
showing the optimal instrument depends on the relative slopes of marginal
benefit and cost. In the capacity market context, the VRR curve's explicit
downward-sloping demand design embeds the regulator's preference for a price
instrument over a fixed quantity mandate --- an instrument choice that is more
informationally efficient when the equilibrium price location is uncertain.
Suppressing that price with a cap undermines the instrument the market design
was built to use. Restoring a 36-month lead time allows prices to elicit entry
before they must be suppressed; capping prices without restoring lead time
removes the instrument before it can function.

PJM's own settlement framing is consistent with this priority ordering.
The settlement slides describe the 2026/27 BRA (July 2025) and 2027/28 BRA
(January 2026) as a partial restoration of the forward-market schedule
\citep{PJM2025_settlement_slides}. The settlement is presented as a temporary
compromise for the two auctions whose schedules were most disrupted by the
2021--2025 irregularities, not as a permanent recalibration of the VRR curve.
The temporary nature of the settlement --- two delivery years, after which the
tariff would revert to the original VRR calibration or a reformed version ---
limits the cumulative informational damage but does not eliminate it.

% ------------------------------------------------------------------------------
\subsection{Summary}
\label{subsec:lt_summary}
% ------------------------------------------------------------------------------

The lead-time argument provides a coherent justification for treating the
2026/27 and 2027/28 settlements as emergency measures during a period of
acute auction-schedule disruption. Its empirical strength depends on the
deployment horizon of the marginal resource: the argument is strongest for
greenfield gas generation (36--48 months) and weakest for demand response,
storage, and existing-plant upgrades (6--18 months). For the 2026/27 BRA
held 10.2~months ahead, prices could have attracted resources in the second
category, and the informational channels enumerated in
Section~\ref{subsec:lt_channels} remain active regardless of lead time.
The correct structural remedy for the lead-time problem is schedule
restoration --- which PJM is pursuing --- rather than price suppression,
which eliminates the scarcity signal that a restored schedule would
otherwise transmit.
