% ==============================================================================
% Section: Capped Auctions — Case Studies and Policy Implications
% Status: REVISED (2026-04-26 — Phase C of cap-incidence pivot)
%   Replaces the SFE-as-counterfactual framing with the IMM unrestricted-VRR
%   counterfactual; promotes existing concession passages to load-bearing
%   prose; absorbs lead-time and structural-alternatives content from
%   previously-separate sec7_leadtime and sec9_policy_alternatives.
% ==============================================================================

\section{Capped Auctions: Case Studies and Policy Implications}
\label{sec:21billion}

The cap-incidence panel of Section~\ref{sec:cap_incidence} and the bunching
diagnostic of Section~\ref{sec:bunching} together establish that the
binding administrative constraint changed across the sample period from
the cost-basis Net ACR cap (slack years) to the Shapiro/PJM settlement
cap (2026/27 and 2027/28), without a corresponding behavioral shift in
seller bidding. This section turns to the dollar consequences. I treat
the two capped auctions as case studies, anchor the counterfactual in
the IMM's published unrestricted-VRR scenario, and discuss the
reconciliation with the \$21~billion consumer-savings claim, lead-time
compression, and structural alternatives to direct price control.

% ------------------------------------------------------------------------------
\subsection{The 2026/27 Auction: Marginal Shortfall, Price Suppression}
\label{subsec:case_2026}
% ------------------------------------------------------------------------------

The 2026/27 BRA cleared at \$329.17/MW-day, which under the new VRR
design embodies the settlement cap. The market cleared 314~MW below the
RTO reliability requirement of 134{,}519.5~MW (a 0.23\% shortfall), and
493~MW below the IRM-implied reserve target; the IMM declared the auction
``not competitive'' \citep{MonitoringAnalytics2025_sotm}. Capacity Performance offers had
unit-specific Market Seller Offer Caps calculated for 6.3\% of
generation resources, and the offer-cap composition (Table~\ref{tab:bunching})
shows roughly half of resources at their default ACR cost-basis cap and
35\% as price-takers.

The IMM's mechanical counterfactual scenario re-clears the same offers
against the unrestricted (pre-settlement) VRR demand curve and yields
total revenues of \$19.29~billion, a 19.7\% increase over the
\$16.12~billion actual revenue \citep{MonitoringAnalytics2025_sotm}.
The counterfactual operates almost entirely through the price channel:
under the unrestricted VRR, the clearing price would have settled
higher on the demand curve while the cleared quantity would have changed
modestly. With reserve margin only marginally negative, the cap
truncated the price signal more than it truncated the quantity
procured.

% ------------------------------------------------------------------------------
\subsection{The 2027/28 Auction: Material Shortfall, Quantity Rationing}
\label{subsec:case_2027}
% ------------------------------------------------------------------------------

The 2027/28 BRA cleared at \$333.44/MW-day, again at the new-design
Point~(a). The market cleared 6{,}623~MW below the RTO reliability
requirement of 141{,}101~MW, a 4.7\% shortfall that the IMM
characterized as material; the auction was declared ``not competitive''
on the same basis as 2026/27. Unit-specific MSOCs were calculated for only
1.0\% of resources, consistent with the settlement cap moving close
enough to the underlying offer distribution that fewer individual
offers exceeded their cost-basis caps.

The IMM unrestricted-VRR counterfactual produces total revenues of
\$26.32~billion, a 60.4\% increase over the \$16.41~billion actual
revenue \citep{MonitoringAnalytics2025_sotm}. The disproportionately
large revenue increase relative to 2026/27 reflects two compounding
channels: a higher counterfactual price as the unrestricted VRR allows
clearing further up the demand curve, and a higher counterfactual
cleared quantity as offers above the settlement cap that were excluded
under the actual auction would have cleared. A back-of-envelope
decomposition: at a counterfactual price of \$500/MW-day and the
actual cleared quantity of 134{,}478~MW over 366 days, revenues would
be approximately \$24.6~billion. The IMM's \$26.32~billion estimate
exceeds this by roughly \$1.7~billion, implying counterfactual cleared
quantity of approximately 144{,}000~MW---9{,}500~MW above actual and
roughly equal to the reliability requirement plus a positive reserve
margin. Under the IMM scenario, in other words, the unrestricted
auction would have procured both higher prices and adequate capacity;
under the cap, the auction procured neither.

The 2027/28 case is thus the cleanest evidence in the sample of the
quantity-rationing channel of a binding administrative cap. In
2026/27 the cap suppressed the price signal but procurement was
nearly adequate; in 2027/28 the cap suppressed prices and excluded
the marginal supply needed to meet the reliability requirement.

% ------------------------------------------------------------------------------
\subsection{Reconciling the \$21 Billion Savings Claim}
\label{subsec:21b_comparison}
% ------------------------------------------------------------------------------

Pennsylvania's January 28, 2025 announcement projected consumer savings
of ``over \$21~billion over the next two years''
\citep{Shapiro2025_press_release}. The arithmetic uses a counterfactual
price ``over \$500/MW-Day'' \citep{Shapiro2025_press_release},
consistent with the pre-settlement VRR Point~(a) price visualized in
PJM's settlement consultation \citep{PJM2025_settlement_slides}, applied
to two delivery years of cleared quantity.

Table~\ref{tab:capped_revenue} compares the actual cleared revenue to
the IMM unrestricted-VRR counterfactual and to a uniform-\$500 price
counterfactual at actual cleared quantities.

\begin{table}[H]
  \centering
  \caption{Capacity Market Revenue Under Counterfactual Cap Regimes
    (System Total)}
  \label{tab:capped_revenue}
  \small
  \begin{tabular}{lcccc}
    \toprule
                                                 & 2026/27 & 2027/28 & Two-year & Implied \\
    Scenario                                     & revenue & revenue & total    & savings \\
    \midrule
    Actual (settlement-modified VRR)             & \$16.12B & \$16.41B & \$32.53B & --- \\
    \addlinespace
    IMM unrestricted-VRR counterfactual          & \$19.29B & \$26.32B & \$45.62B & \$13.08B \\
    \quad \textit{Percent increase vs.\ actual}  & +19.7\%  & +60.4\%  & +40.2\%  & --- \\
    \addlinespace
    \$500/MW-day at actual cleared MW            & \$24.51B & \$24.60B & \$49.10B & \$16.57B \\
    \addlinespace
    Stated Pennsylvania claim                    & ---      & ---      & $\sim$\$21B & $\sim$\$21B \\
    \bottomrule
  \end{tabular}
  \vspace{3pt}

  \begin{minipage}{0.96\textwidth}
    \footnotesize \textit{Notes:} Revenues in nominal dollars over the
    twelve-month delivery year (366 days for 2027/28). The IMM
    unrestricted-VRR counterfactual holds offers fixed and re-clears
    against the pre-settlement VRR demand curve; reported figures are
    from the IMM 2025 State of the Market Report
    \citep{MonitoringAnalytics2025_sotm}. The \$500/MW-day scenario
    multiplies the actual cleared UCAP (134{,}205~MW in 2026/27,
    134{,}478~MW in 2027/28) by Pennsylvania's stated counterfactual
    price; this scenario corresponds to the Pennsylvania-stated
    counterfactual but holds quantity at the capped level rather than
    re-clearing demand. The \$21~billion claim
    \citep{Shapiro2025_press_release} reflects additional revenue
    contributions from constrained Locational Deliverability Areas
    where local Point~(a) prices substantially exceeded \$325/MW-day in
    prior years. Savings figures are differences between the
    counterfactual scenario and the actual outcome.
  \end{minipage}
\end{table}

The IMM unrestricted-VRR counterfactual implies two-year savings of
\$13.08~billion at the system level: about three-fifths of the
\$21~billion figure but more than seven-tenths of the \$16.57~billion
implied by Shapiro's counterfactual price applied to actual cleared
MW. Three observations follow.

First, the IMM counterfactual is not a full behavioral counterfactual.
It re-clears with offers held fixed against an alternative demand
curve, which captures the mechanical effect of restoring the
unrestricted VRR but not the strategic re-bidding sellers would
undertake in a no-cap world. The behavioral counterfactual would
typically yield a clearing price below the IMM mechanical estimate,
because in the unrestricted scenario supply tightness loosens (the
unrestricted VRR procures more capacity), and a less-tight market
clears at lower equilibrium prices for any given level of seller
strategy.

Second, the gap between the IMM \$13.08~billion and the stated
\$21~billion is partly accounted for by constrained-LDA effects that
the system-level IMM scenario does not separately enumerate. Constrained
LDAs have local Point~(a) prices substantially above the system-level
value, so revenue increases under an unrestricted scenario are
correspondingly larger.

Third, the structural significance of either savings figure depends on
how the savings are interpreted. If they represent rents transferred
from sellers to consumers in years when scarcity is real, they reduce
the incentive for entry and retirement deferral that the capacity
market is designed to provide. If they represent rents that would have
gone to sellers exercising market power, they are unambiguously
welfare-improving. Section~\ref{sec:cap_incidence} argues that the
binding constraint in 2026/27 and 2027/28 was the settlement cap
overlaid on a market entering scarcity, not a strategic-conduct
phenomenon driven by seller behavior; under that interpretation the
savings come at the cost of a truncated investment signal during the
years it would have been most informative.

The truncated investment signal interacts with the compressed lead
times of these auctions. The 2025/26 and 2026/27 BRAs were held only
about 10 months before the start of the delivery year and the 2027/28
BRA 17.7 months ahead (Table~\ref{tab:bra_timeline}), against a
canonical three-year forward design. A 10-month horizon is too short
for greenfield gas but non-trivial for demand response, battery storage,
and capacity upgrades to existing plants, and auction prices inform
retirement deferral on the same timescale \citep{Cramton2005_capacity}.
A two-year cap during a tight-supply episode therefore disrupts the
retirement and short-deployment signal precisely in the years when it
would have been most informative.

