% ==============================================================================
% Section: Policy Alternatives
% Status: COMPLETE (2026-03-28 Session R5)
% ==============================================================================

\section{Policy Alternatives}
\label{sec:policy_alternatives}

The analysis in Sections~\ref{sec:allocative_cost}--\ref{sec:21billion}
establishes that the Shapiro settlement reduces the equilibrium capacity price
by \$4--\$8/MW-day at the RTO level, suppresses the scarcity signal that
informs entry decisions, and generates aggregate consumer savings that are an
order of magnitude smaller than the VRR-counterfactual estimate of
\$21~billion. This section considers four structural alternatives that address
the source of elevated capacity prices --- a concentrated seller side with
limited competitive fringe --- rather than suppressing the price symptom.
The alternatives are not mutually exclusive; the most durable solution likely
combines elements of all four.

% ------------------------------------------------------------------------------
\subsection{Option 1: The Shapiro Settlement (Current Policy)}
\label{subsec:pol_settlement}
% ------------------------------------------------------------------------------

The settlement reshapes the VRR demand curve for 2026/27 and 2027/28, imposing
a cap at $\bar{p}_S = \$325$/MW-day UCAP and a floor at $\bar{p}_f = \$175$/MW-day
UCAP. It is designed as a temporary emergency measure during the period of
acute auction-schedule disruption documented in Section~\ref{subsec:inst_leadtime},
and PJM's slides describe it as applicable to two specific delivery years
only \citep{PJM2025_settlement_slides}.

The settlement's key advantages are speed and certainty: the VRR curve
modification is directly within FERC's regulatory authority under FPA~\S205,
it does not require structural changes to the seller side, and it provides
immediate price relief for the 2026/27 and 2027/28 auctions. Its key
disadvantages are equally clear. It does not reduce the number of strategic
sellers, expand the competitive fringe, or steepen the VRR demand curve;
it therefore does nothing to address the structural source of the $K \leq 3$
regime. Section~\ref{sec:allocative_cost} shows that the price-cost margins
in 2026/27 would remain at $L = 0.54$ under the settlement --- unchanged
from the SFE equilibrium. The settlement compresses the \emph{price}, but
not the underlying \emph{markup structure}.

Furthermore, as Section~\ref{sec:leadtime} argues, suppressing the price
signal imposes an informational cost that persists beyond the settlement
window: investors planning for 2028/29 and later years observe capped prices
rather than the unconstrained equilibrium, reducing the clarity of the
capacity market's investment signal.

% ------------------------------------------------------------------------------
\subsection{Option 2: Transmission Expansion (CETL)}
\label{subsec:pol_cetl}
% ------------------------------------------------------------------------------

The comparative statics in Section~\ref{subsec:res_transmission} show that
increasing the competitive fringe supply $\bar{Q}_f$ reduces equilibrium
prices organically, without requiring any modification to the demand curve.
Fringe supply in the SFE model represents capacity that bids at (or near)
marginal cost, which in PJM's context corresponds to imports limited by
the Capacity Emergency Transfer Limit. A 30\% increase in the competitive
fringe at $K = 3$ reduces the price-cost margin from 0.54 to approximately
0.30 in 2026/27 --- a reduction of 24~percentage points, far larger than
the settlement's zero-point-zero reduction in price-cost margins.

The LDA cross-sectional results (Section~\ref{subsec:res_lda}) confirm this
mechanism: a 10-percentage-point increase in the import penetration ratio
$\text{CETL}/\text{reliability requirement}$ is associated with a 4--6-point
reduction in the LDA price-cost margin. For the median LDA in the sample
(import penetration 0.40), increasing the CETL to achieve a 0.50 import
penetration ratio would reduce the price-cost margin by roughly 6--12 points,
equivalent to a price reduction of \$20--\$40/MW-day at the 2026/27 VRR cap.

CETL expansion through transmission investment preserves the price signal:
equilibrium prices fall because competition increases, not because the demand
curve is truncated. The scarcity signal remains informative; it is simply
responding to a market structure with more competitive imports. This makes
transmission expansion the most structurally efficient of the four options
considered here. The relevant policy instrument is PJM's Regional Transmission
Expansion Plan (RTEP), which can be directed to prioritize projects that
increase import capacity to capacity-constrained LDAs.

% ------------------------------------------------------------------------------
\subsection{Option 3: VRR Curve Redesign}
\label{subsec:pol_vrr}
% ------------------------------------------------------------------------------

Section~\ref{subsec:res_vrr} shows that steepening the VRR demand curve ---
increasing $|D'(p)|$ on the sloped segment --- reduces equilibrium markups
through the ODE mechanism. The intuition is straightforward: a steeper demand
curve means each unit of capacity withheld has less effect on price (the
curve is harder to move), which reduces the marginal incentive to withhold.
A doubling of the VRR slope reduces the price-cost margin by 20--30
percentage points in interior-clearing years; for 2026/27, the at-cap outcome
transitions to interior clearing at a slope scaling factor of approximately
1.4.

FERC Order ER25-1357 redesigned the VRR curve for the 2026/27 BRA, steepening
the sloped segment between $p_f$ and $\bar{p}$ by narrowing its quantity width.
This is a step in the right direction from an SFE perspective. However, the
redesign also introduced the flat floor at $p_f \approx \$177$/MW-day, which
as Section~\ref{subsec:alloc_floor} discusses, creates an investment-adequacy
guarantee whose interaction with the cap introduces its own distortions.

A reformed VRR curve that steepens the slope further while maintaining a
smooth lower tail (without a hard floor) would more directly address the
strategic markup problem. The quantitative target from the SFE results is
a slope factor of approximately 1.4 or greater: at that steepness, the
2026/27 equilibrium transitions from cap-binding to interior-clearing even
with $K = 3$ sellers, allowing the price signal to operate freely below the
Point~(a) cap.

% ------------------------------------------------------------------------------
\subsection{Option 4: Entry Promotion and Deconcentration}
\label{subsec:pol_entry}
% ------------------------------------------------------------------------------

The $K$ comparative static in Section~\ref{subsec:res_concentration} is the
most direct guide to durable structural reform. The transition from
cap-binding to interior-clearing equilibria occurs between $K = 3$ and
$K = 4$: adding a fourth symmetric strategic seller reduces the price-cost
margin by 20--33 percentage points and moves the 2026/27 equilibrium from
\$329 to \$228 (a reduction of \$101/MW-day). With $K \geq 7$, equilibrium
prices fall to within 0.5\% of marginal cost in all three benchmark years.

Reducing effective seller concentration in PJM's capacity market requires
lowering the barriers that prevent entry and expansion. Three levers are
available:

\begin{enumerate}

\item \textbf{Interconnection queue reform.} PJM's interconnection queue has
historically taken 4--7 years to process new generator applications, far
exceeding the lead times at which high capacity prices can stimulate entry.
FERC Order 2023 on transmission planning and Order 2023-A on interconnection
reform are designed to accelerate this process. Timely interconnection is
a necessary condition for the $K \geq 4$ outcome that would organically
resolve the cap-binding equilibrium.

\item \textbf{Demand response and storage deployment.} Demand response
resources and utility-scale battery storage can participate in PJM's
capacity market on timescales much shorter than new thermal generation.
Expanding the effective capacity of these resources increases $K$ in the
SFE framework by adding sellers who do not share the symmetric cost
structure of the thermal strategic sellers. This is an asymmetric effect:
demand response that bids at zero marginal cost is analogous to increasing
the competitive fringe rather than increasing $K$, providing the fringe-expansion
benefits described in Option~2.

\item \textbf{Targeted divestiture or mitigation.} For the LDAs at the left
tail of the import-penetration distribution --- PS, PS~NORTH, COMED, and
EMAAC --- even large CETL expansions leave price-cost margins above 0.50
given the current seller count (Section~\ref{subsec:res_lda}). In these
areas, structural remedies including divestiture or enhanced offer price
mitigation may be necessary to achieve interior clearing. PJM's existing
Capacity Market Seller Offer Cap mitigation rules provide a framework for
this, though they have historically been calibrated to the Avoidable Cost
Rate rather than the SFE-based markup threshold.

\end{enumerate}

% ------------------------------------------------------------------------------
\subsection{Comparative Assessment}
\label{subsec:pol_ranking}
% ------------------------------------------------------------------------------

Table~\ref{tab:policy_comparison} summarizes the four options along the
dimensions that matter for capacity market design: the source of price
reduction, the duration of the effect, whether the price signal is
preserved, and the required regulatory authority.

\begin{table}[ht]
  \centering
  \caption{Comparative Assessment of Policy Alternatives}
  \label{tab:policy_comparison}
  \begin{tabular}{p{3.2cm}p{2.8cm}lp{2cm}p{3cm}}
    \hline\hline
    Option & Price reduction mechanism & Duration & Signal preserved & Regulatory lever \\
    \hline
    1. Settlement (current) & Cap/floor reshape VRR & 2 years & No (cap binding) & FERC \S205 tariff \\
    2. CETL expansion       & Competitive fringe $\uparrow$ & Permanent & Yes & RTEP, FERC Tx planning \\
    3. VRR redesign         & Demand elasticity $\uparrow$ & Permanent & Yes & FERC VRR recalibration \\
    4. Entry/deconcentration & $K \uparrow$ / fringe $\uparrow$ & Permanent & Yes & FERC Ords. 2023/2023-A; mitigation \\
    \hline\hline
  \end{tabular}
  \smallskip

  \noindent\footnotesize
  \textit{Notes:} Signal preservation indicates whether the capacity clearing
  price reflects unconstrained strategic equilibrium (Options~2--4) or is
  truncated by an administrative cap (Option~1). Duration refers to whether
  the price reduction persists beyond a fixed delivery window.
\end{table}

Options~2--4 are strictly preferable to Option~1 along every structural
dimension: they produce permanent price reductions by addressing the source
of high prices (concentrated supply, limited competitive fringe, inelastic
demand) rather than the symptom (high clearing prices). The settlement is
justified as an emergency measure during a period of acute schedule
disruption and lead-time compression, but it should not substitute for the
structural reforms that would render a similar intervention unnecessary in
future auctions. PJM's own settlement framing calls for ``reform the
capacity market'' alongside the cap \citep{PJM2025_settlement_slides},
consistent with this ranking.
