\documentclass[12pt]{article}

% Preamble (loaded from Preambles/ via TEXINPUTS)
\input{header}

% ==============================================================================
\begin{document}
% ==============================================================================

% --- Title page ---
\title{Capping the Capacity Market: A Supply Function Equilibrium Analysis of Price Controls in PJM}
\author{}
\date{\today}
\maketitle

\begin{abstract}
In January 2025, Pennsylvania Governor Josh Shapiro negotiated a settlement
reshaping PJM's capacity market demand curve: a price cap at \$325/MW-day
UCAP and a price floor at \$175/MW-day UCAP for the 2026/27 and 2027/28
delivery years, projected to save consumers over \$21~billion. We evaluate
this intervention using a calibrated Supply Function Equilibrium model of
PJM's Base Residual Auction. The settlement cap is binding: under the
observed market structure with three collectively pivotal suppliers, the
model predicts clearing prices at or above the VRR Point~(a) cap
(\$329/MW-day in 2026/27), which exceeds the settlement cap by
\$4--\$8/MW-day at the RTO level, with larger gaps in constrained
Locational Deliverability Areas. The settlement therefore suppresses
prices but leaves the underlying price-cost margin nearly unchanged. The
\$21~billion savings figure uses the unconstrained VRR Point~(a) price
as the counterfactual; a structural SFE benchmark implies a substantially
smaller two-year revenue transfer, because the equilibrium under the model's
assumptions falls near the design ceiling rather than far above the
settlement cap. Critically, the \$21~billion figure also holds cleared
quantities fixed, overstating savings to the extent the cap depresses future
entry and accelerates retirement. The complaint's lead-time argument is
partial: even at short auction horizons, prices inform retirement deterrence,
mobile resource deployment, contracting, and forward investment expectations.
Transmission expansion (increasing the Capacity Emergency Transfer Limit)
reduces equilibrium prices organically, preserving the scarcity signal while
addressing the structural source of binding settlement outcomes.
\end{abstract}

\newpage
\tableofcontents
\newpage

% ==============================================================================
% Sections
% ==============================================================================

\input{sections/introduction}

\input{sections/literature}

\input{sections/institutional}

\input{sections/model}

\input{sections/calibration}

\input{sections/results}

\input{sections/sec6_allocative_cost}

\input{sections/sec7_leadtime}

\input{sections/sec8_21billion}

\input{sections/sec9_policy_alternatives}

\input{sections/conclusion}

% ==============================================================================
% Bibliography
% ==============================================================================

\newpage
\bibliography{../Bibliography_base}

% ==============================================================================
% Appendix (future)
% ==============================================================================
% \begin{appendices}
%   \input{sections/appendix_proofs}
%   \input{sections/appendix_data}
% \end{appendices}

\end{document}
